# Working with **ggplot2**

First, the **ggplot2** package comes with the **tidyverse** meta-package. 
You can just load **tidyverse**, and it will also load **ggplot2**.

Let's use the above scatterplot as an example to see how to use the `ggplot()` function.

First, we will pass one argument to `ggplot()` function itself:
  - `data`: the data frame containing your plotting data
  
```{r}
ggplot(gapminder)
```
  
After this, we **add** additional functions as **layers** to add features to the plot and control their appearance. The first thing we will add is `aes()`. The `aes()` function tells R to look in the `data` for variable names to *map* them to different *aesthetic features*, such as x-position, y-position, or color.

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp)
```

Next, we want to add some **geometric shapes** to the plot. 
These use functions with the form `geom_SOMETHING()`.
Different plot shapes (`geom_SOMETHING`) accept different `aes()` arguments.
These are listed in the help file for the geom function: 
`?geom_point`, `?geom_line`, etc.

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp) + 
  geom_point()
```

You can control aesthetics using constants or variables from outside `data` by specifying them outside of `aes()`. 
For example, to make all of the shapes of plot blue, you can add: `color = "blue"`.

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp) + 
  geom_point(color = "blue")
```

There's a bit of overplotting (overlapping symbols), so let's also make the points semi-transparent.
This is controlled using the `alpha` argument 
(you need `1/alpha` points overlaid to achieve a solid point).

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp) + 
  geom_point(color = "blue", alpha = .1)
```

For now, that's the only `geom` that we want to add. 
Now, let's specify a scale transformation, because the plot would really benefit if the x-axis is on a logarithmic 
scale. 
These functions take the form `scale_AESTHETIC_TYPE()`. 

As usual, you can tweak this layer, too, using this function's arguments. 
In this example, we're re-naming the x-axis (the `name` argument), 
transform the values (the `trans` argument), 
and changing the labels to have a dollar format (the `labels` argument).

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp) + 
  geom_point(color = "blue", alpha = .1) +
  scale_x_continuous(
    name = "GDP per capita", 
    trans = "log10", 
    labels = scales::dollar_format()
  )
```

I'm tired of seeing the ugly default gray background, so I'll add a `theme()` 
layer. 
I like `theme_bw()` (you can tweak themes later, too!). 
Then, I'll re-label the y-axis using the `ylab()` function. 
Et voilà!

```{r}
ggplot(gapminder) +
  aes(x = gdpPercap, y = lifeExp) + 
  geom_point(color = "blue", alpha = .1) +
  scale_x_continuous(
    name = "GDP per capita", 
    trans = "log10", 
    labels = scales::dollar_format()
  ) +
  theme_bw() +
  ylab("Life expectancy")
```
