# Activity: Build some plots!

## Make a Line Chart

The following code makes a data frame called `mauna` that contains time series data of CO$_2$ concentrations collected monthly at the Mauna Loa vocanic observation station.

Execute this code to store the data in `mauna`:

```{r}
(mauna <- tsibble::as_tsibble(co2) |> 
   rename(month = index, conc = value))
```

Produce a line chart showing the concentration over time. 
Specifically, the plot should have the following grammar components:

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `mauna` |
| __aesthetic mapping__ | x: month, y: conc |
| __geometric object__  | lines |
| scale                 | yearmonth |
| statistical transform | none |
| coordinate system     | rectangular |
| faceting              | none |

Fill in the blanks to obtain the plot:

```{r, fig.width = 5, fig.height = 2, eval = FALSE}
ggplot(FILL_THIS_IN) +
  aes(FILL_THIS_IN, FILL_THIS_IN)
  FILL_THIS_IN() + 
  tsibble::scale_x_yearmonth()
```

## Make a Scatterplot

Use the `palmerpenguins::penguins` data to make a scatterplot with the following specifications:

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `palmerpenguins::penguins` |
| __aesthetic mapping__ | x: body_mass_g, y: bill_depth_mm |
| __geometric object__  | points, smoothed lines |
| scale                 | linear |
| statistical transform | none |
| coordinate system     | rectangular |
| faceting              | none |


```{r, fig.width = 5, fig.height = 2, eval = FALSE}
ggplot(FILL_THIS_IN) +
  aes(FILL_THIS_IN, FILL_THIS_IN)
  FILL_THIS_IN() + 
  geom_smooth()
```

You can control aesthetics for individual layers by adding an `aes()` inside the layer function, like this:

```
geom_point(aes(color = COLOR_VARIABLE))
````

Modify your code above so that the points (but not the smooth line) have their `color` mapped to `species`:

```{r, fig.width = 5, fig.height = 2, eval = FALSE}
ggplot(FILL_THIS_IN) +
  aes(FILL_THIS_IN, FILL_THIS_IN)
  FILL_THIS_IN() + 
  geom_smooth()
```

Now, instead, map `color` in the global `aes()` call for the plot.
What happens?

```{r, fig.width = 5, fig.height = 2, eval = FALSE}
ggplot(FILL_THIS_IN) +
  aes(FILL_THIS_IN, FILL_THIS_IN)
  FILL_THIS_IN() + 
  geom_smooth()
```

Things to keep in mind:

- Aesthetics mapped outside of a specific layer apply **globally**
- Aesthetics mapped inside a `geom` layer apply only to that layer.
- If the same aesthetic appears both globally and in a layer, the layer-specific one wins


## Assigning a **ggplot2** object and adding to it

You can store the output of the plot in a variable. 
Assign the `mauan` plot above to a variable named `p`, 
then add a layer to `p` that adds a dark green smoothed line to the plot.

```{r, fig.width = 5, fig.height = 2, eval = FALSE}
FILL_THIS_IN

p +
  FILL_THIS_IN()
```


## Fix Me!

What's wrong with the following code? Fix it.

```{r, fig.width = 5, fig.height = 2, eval = FALSE}
ggplot(gapminder) +
  geom_point(x = gdpPercap, y = lifeExp, alpha = 0.1)
```

What's wrong with this code? Fix it.

```{r, fig.width = 5, fig.height = 2, eval = FALSE}
ggplot(cars) +
  geom_point(aes(x = speed, y = dist, color = "blue"))
```
