# Grouping and summarizing data

## `summarize()`

Like `mutate()`, the `summarize()` function also creates new columns, 
but the calculations that make the new columns must reduce down to a single number. 

For example, let's compute the mean and standard deviation of life expectancy 
in the `gapminder` data set:

```{r}
gapminder |> 
  summarize(
    mean = mean(lifeExp),
    sd = sd(lifeExp)
  )
```

Notice that all other columns were dropped. 
This is necessary, because there's no obvious way to compress the other columns down to a single row. 
This is unlike `mutate()`, which keeps all columns, 
and more like `transmute()`, which drops all other columns.

As it is, this is useful for creating summary tables, 
but it's more useful in the context of _grouping_, coming up next.


## `group_by()`

The true power of **dplyr** lies in its ability to group a tibble, with the 
`group_by()` function. As usual, this function takes in a tibble and returns a 
(grouped) tibble. 

Let's group the gapminder dataset by continent and year:

```{r}
gapminder |> 
  group_by(continent, year)
```

The only thing different from a regular tibble is the indication of grouping 
variables above the tibble. This means that the tibble is recognized as having 
"chunks" defined by unique combinations of continent and year:

- Asia in 1952 is one chunk.
- Asia in 1957 is another chunk.
- Europe in 1952 is another chunk.
- etc...

Notice that the data frame isn't _rearranged_ by chunk! The grouping is something
stored internally about the grouped tibble.

Now that the tibble is grouped, operations that you do on a grouped tibble 
_will be done independently within each chunk_, as if no other chunks exist. 

You can also create new variables and group by that variable simultaneously. 
Try splitting life expectancy by "small" and "large" using 60 as a threshold:

```{r}
gapminder |> 
  group_by(smallLifeExp = lifeExp < 60)
```


## Grouped `summarize()`

Want to compute the mean and standard deviation for each year for every 
continent? No problem:

```{r}
gapminder |> 
  group_by(continent, year) |> 
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))
```

Notice:

- The grouping variables are kept in the tibble, because their values are unique 
  within each chunk (by definition of the chunk!)
- With each call to `summarize()`, the grouping variables are "peeled back" from 
  last grouping variable to first.

This means the above tibble is now only grouped by continent. What happens when 
we reverse the grouping?

```{r}
gapminder |> 
  group_by(year, continent) |>    # Different order
  summarize(mu    = mean(lifeExp),
            sigma = sd(lifeExp))
```

The grouping columns are switched, and now the tibble is grouped by year instead 
of continent. 

**dplyr** has a bunch of convenience functions that help us write code more 
eloquently. We could use `group_by()` and `summarize()` with `length()` to find 
the number of entries each country has:

```{r}
gapminder |> 
  group_by(country) |> 
  transmute(n = length(country))
```

Or, we can use the more elegant `dplyr::n()` to count the number of rows in each 
group:

```{r}
gapminder |> 
  group_by(country) |> 
  summarize(n = n())
```

Or better yet, if this is all we want, just use `dplyr::count()`:

```{r}
gapminder |> 
  count(country)
```


## Grouped `mutate()`

Want to get the increase in GDP per capita for each country? No problem:

```{r}
gap_inc <- gapminder |> 
  arrange(year) |> 
  group_by(country) |>
  mutate(gdpPercap_inc = gdpPercap - lag(gdpPercap))
print(gap_inc)
```

The tibble is still grouped by country.

Drop the `NA`s with another convenience function, this time supplied by the 
`tidyr` package (another tidyverse package that we'll see soon):

```{r}
gap_inc |> 
  tidyr::drop_na()
```

You can specify specific columns to drop `NA`s from in the `drop_na()` function.


## Function types

We've seen cases of transforming variables using `mutate()` and `summarize()`, 
both with and without `group_by()`. How can you know what combination to use? 
Here's a summary based on one of three types of functions.


| Function type | Explanation | Examples | In **dplyr** |
|------|-----|----|----|
| Vectorized functions | These take a *vector*, and operate on each component independently to return a vector of the same length. In other words, they work element-wise. | `cos()`, `sin()`, `log()`, `exp()`, `round()` | `mutate()` |
| Aggregate functions | These take a vector, and return a vector of length 1 | `mean()`, `sd()`, `length()` | `summarize()`, esp with `group_by()`. |
| Window Functions | these take a vector, and return a vector of the same length that depends on the vector as a whole. | `lag()`, `rank()`, `cumsum()` | `mutate()`, esp with `group_by()` |
