# Pivoting data

The task of making tidy data is about making data either _longer_, by stacking two or more rows, or _wider_, by putting one or more columns alongside each other based on groups. 
This is called __pivoting__.

Usually the task of tidying data involves lengthening, and usually the task of widening is useful for turning data into something more friendly for human eyes.

Sometimes, you will see data described as being in "long" or "wide" formats.
Those terms aren't that useful—"long" and "wide" are relative terms. 

The easiest and most powerful way to widen or lengthen data are with the functions `tidyr::pivot_wider()` and `tidyr::pivot_longer()`.

History: R has seen many attempts at reshaping that have progressively gotten better. 

First came the `reshape` and `reshape2` packages. 

  - Both were finicky. 
  - Used function names that I could never remember: `melt()` and `cast()`. 
  
Then, `tidyr` package replaced these.

  - The `tidyr::spread()` and `tidyr::gather()` functions provided a simple interface.
  - I still couldn't remember these names. 
  
We will use `tidyr::pivot_longer()` and `tidyr::pivot_wider()`.

The "pivot" functions also have similar names to the SQL "PIVOT" and "UNPIVOT" functions.


## Univariate pivoting

Let's start with pivoting in the simplest case where only one variable is "out of place". 
We'll use the hair and eye color example from before, using the untidy data version from Example 1:

```{r}
haireye_untidy <- haireye |> 
  mutate(eye = str_c(eye, "_eyed")) |> 
  pivot_wider(id_cols = hair, names_from = eye, values_from = n)

haireye_untidy
```

The _eye color_ variable is spread out across columns. 
To fix this, we need to convert the eye color columns to two columns:

- one column to hold the eye color (column names),
- one column to hold the values.

Doing this, we obtain:

```{r, echo = FALSE}
haireye_untidy |> 
  pivot_longer(
    contains("eyed"), 
    names_to = "eye", 
    values_to = "n"
  )
```

Let's dig into that!

### `pivot_longer()`

`pivot_longer()` takes a data frame, and returns a data frame. 

The main arguments after the data argument that we'll need are:

- `cols`: the columns we want to pivot into a single column. Give the column names
- `names_to`: the old column _names_ are going to be stored in a new column. 
  What should this new column be named?
- `values_to`: the _values_ in the old columns are going to stored in a new column.
  What should this new column be named?

Possibly the trickiest bit is in identifying the column names. 
We could list all of them:

```{r}
haireye_untidy |> 
  pivot_longer(cols      = c(Blue_eyed, Brown_eyed, Green_eyed, Hazel_eyed),
               names_to  = "eye",
               values_to = "n")
```

That can be a little tedious.

We could identify a range. This is efficient, but not so robust if the data changes.

```{r}
haireye_untidy |> 
  pivot_longer(cols      = Blue_eyed:Hazel_eyed,
               names_to  = "eye",
               values_to = "n")
```

Better is to use helper functions from the `tidyselect` package. 
In this case, we know the columns contain the text "eyed", so let's use `tidyselect::contains()`:

```{r}
haireye_untidy |> 
  pivot_longer(cols      = contains("eyed"),
               names_to  = "eye",
               values_to = "n")
```

Yet another way is to indicate everything except the `hair` column:

```{r}
haireye_untidy |> 
  pivot_longer(cols      = -hair,
               names_to  = "eye",
               values_to = "n")
```


### `pivot_wider()`

Let's say we want go from a longer data frame to a shorter data frame.
This is the opposite of what we did above. 

We might want to do this:

  - To make a table for presentation
  - With longitudinal or family data, to to go from **multilevel models** (which need longer data; each row is an *observation*) to **SEM analyses** (which need wider data; each row is an *individual* or *family*)
  
For example, if we want to go from:

```{r, echo = FALSE}
haireye_untidy |> 
  pivot_longer(
    contains("eyed"), 
    names_to = "eye", 
    values_to = "n"
  )
```

To: 

```{r, echo = FALSE}
haireye_untidy
```

We need to: 

  - Take the column `eye` and make *each unique entry* a new column
  - Take the column `n` and make these values in the new `eye` columns.

`pivot_wider()` is the reverse of `pivot_longer()`.

Like `pivot_longer()`, `pivot_wider()` takes a data frame and returns a data frame. 
The main arguments after the data argument that we'll need are:

- `id_cols`: The columns you would like to **keep**. 
  For example, the identifier number for the observation.
- `names_from`: The new column names are coming from an old column. 
  Which column is this?
- `values_from`: The column values are coming from an old column. 
  Which column is this?

```{r}
haireye |> 
  pivot_wider(id_cols     = hair, 
              names_from  = eye, 
              values_from = n)
```
