# Multivariate pivoting

Now let's consider the case when more than one variable are "out of place".
Perhaps there are multiple variables per row and/or multiple observations per row.


## Multiple variables in column names

Consider this subset of the `who` data:

```{r, echo = TRUE}
WHO <- who |>
    select(country:year, starts_with("new_")) |> 
    rename_with(~ stringr::str_replace(.x, "f", "f_"), starts_with("new_")) |> 
    rename_with(~ stringr::str_replace(.x, "m", "m_"), starts_with("new_"))
knitr::kable(WHO, rownames = FALSE)
```

`country`, `iso2`, `iso3`, and `year` are already variables, so they can be left as is. 
But the columns from `new_sp_m_014` to `new_ep_f_65` encode four variables in their names:

* The `new` prefix indicates these are counts of new cases (versus total cases). 
  This dataset only contains new cases, so we'll ignore it here because it's constant.
  
* `sp`/`rel`/`ep` describe how the case was diagnosed.

* `m`/`f` gives the gender.

* `014`/`1524`/`2535`/`3544`/`4554`/`65` supplies the age range.

We can break these variables up by specifying multiple column names in `names_to`, and then providing `names_sep`.

```{r}
WHO |> pivot_longer(
  cols = new_sp_m_014:new_ep_f_65,
  names_to = c("diagnosis", "gender", "age"), 
  names_prefix = "new_",
  names_sep = "_",
  values_to = "count"
)
```

In the `names_to` argument, we now tell it the names of the new columns that will store each part of the existing column names.
We give the column names in order corresponding to how they appear in the existing column names.

As with univariate pivoting, `values_to` gives the name of the new column that will store the cell values.

## Multiple variables in column names

Consider these family data. 

```{r}
fam_dat <- tribble(
  ~family,  ~dob_child1,  ~dob_child2, ~gender_child1, ~gender_child2,
       1L, "1998-11-26", "2000-01-29",             1L,             2L,
       2L, "1996-06-22",           NA,             2L,             NA,
       3L, "2002-07-11", "2004-04-05",             2L,             2L,
       4L, "2004-10-10", "2009-08-27",             1L,             1L,
       5L, "2000-12-05", "2005-02-28",             2L,             1L,
)
fam_dat <- fam_dat |> mutate_at(vars(starts_with("dob")), parse_date)
fam_dat
```

In these data, we have two pieces of information (or values) for each child: their `gender` and their `dob` (date of birth). 
These need to go into separate columns in the result. 
Again we supply multiple variables to `names_to`, using `names_sep` to split up each variable name. 

The `names_to` vector gives the names of the new columns that will store each part of the existing column names.
We give the column names in order corresponding to how they appear in the existing column names.

Note the special name `.value`: 

`.value` takes the place of the `values_to` argument.
It tells `pivot_longer()` to get the name of the column that will hold the cell values from that part of the existing column name.

```{r}
fam_dat |> 
  pivot_longer(
    cols = -family,
    names_to = c(".value", "child"), 
    names_sep = "_", 
    values_drop_na = TRUE
  )
```

Let's also clean up the `child` column:

```{r}
fam_dat_long <- fam_dat |> 
  pivot_longer(
    cols = -family,
    names_to = c(".value", "child"), 
    names_sep = "_", 
    values_drop_na = TRUE
  ) |> 
  mutate(child = stringr::str_replace(child, "child", "")) |> 
  mutate(child = as.integer(child))
```


## `pivot_wider()`

You can also `pivot_wider()` while using multiple columns to supply variable names:

- `id_cols`: as usual.
- `names_from`: the new variable names are coming from old columns. Which old columns?
- `names_sep`: What character should you separate the entries of the old columns by?
- `values_from`: as usual.

```{r}
fam_dat_long |> 
  pivot_wider(id_cols     = family,
              names_from  = c(child, gender), 
              names_prefix = "child",
              names_sep   = "_gender", 
              values_from = dob)
```

Or using multiple columns to supply new values:

If variables are spread out amongst rows _and_ columns (for example, "sepal width" has "sepal" in a column, and "width" as a column name), here's how we can use `pivot_wider()`:

- `id_cols`: as usual
- `names_from`: Which column contains the part of the variable?
- `names_sep`: As before, what character should you separate the entries of the old columns by?
- `values_from`: Which column names contain the other part of the variable?

```{r}
fam_dat_long |> 
  pivot_wider(id_cols     = family, 
              names_from  = child,
              names_prefix = "child",
              names_sep   = "_",
              values_from = c(dob, gender))
```
