# Lab 06: Join those tables!

Load required packages:

```{r, echo = TRUE}
library(tidyverse)
```

## Exercise 1: `singer`

The package `singer` comes with two smallish data frames about songs. 
Let's take a look at them (after minor modifications by renaming and shuffling):

You can download the singer data from the class repo:

```{r, eval = FALSE}
songs <- read_csv("https://raw.githubusercontent.com/bwiernik/progdata-class/master/data/singer/songs.csv")
locations <- read_csv("https://raw.githubusercontent.com/bwiernik/progdata-class/master/data/singer/loc.csv")

(time <- as_tibble(songs) |> 
   rename(song = title))

(album <- as_tibble(locations) |> 
   select(title, everything()) |> 
   rename(album = release,
          song  = title))
```


1. We really care about the songs in `time`. But, for which of those songs do we 
   know its corresponding album?

```{r, eval = FALSE}
time |> 
  FILL_THIS_IN(album, by = FILL_THIS_IN)
```

2. Go ahead and add the corresponding albums to the `time` tibble, being sure to 
   preserve rows even if album info is not readily available.

```{r, eval = FALSE}
time |> 
  FILL_THIS_IN(album, by = FILL_THIS_IN)
```

3. Which songs do we have "year", but not album info?

```{r, eval = FALSE}
time |> 
  FILL_THIS_IN(album, by = "song")
```

4. Which artists are in `time`, but not in `album`?

```{r, eval = FALSE}
time |> 
  anti_join(album, by = "FILL_THIS_IN")
```


5. You've come across these two tibbles, and just wish all the info was 
   available in one tibble. What would you do?

```{r, eval = FALSE}
FILL_THIS_IN |> 
  FILL_THIS_IN(FILL_THIS_IN, by = "song")
```


## Exercise 2: LOTR

Load in three tibbles of data on the Lord of the Rings:

```{r, eval = FALSE}
fell <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv")
ttow <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv")
retk <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Return_Of_The_King.csv")
```

1. Stack these into a single tibble.

```{r}
FILL_THIS_IN(fell, FILL_THIS_IN)
```

2. Which races are present in "The Fellowship of the Ring" (`fell`), but not in 
   any of the other ones?

```{r}
fell |> 
  FILL_THIS_IN(FILL_THIS_IN, by = "Race") |> 
  FILL_THIS_IN(FILL_THIS_IN, by = "Race")
```


## Exercise 3: Set Operations

Let's use three set functions: `intersect`, `union` and `setdiff`. We'll work 
with two toy tibbles named `y` and `z`, similar to Data Wrangling Cheatsheet

```{r}
(y <-  tibble(x1 = LETTERS[1:3], x2 = 1:3))
```

```{r}
(z <- tibble(x1 = c("B", "C", "D"), x2 = 2:4))
```

1. Rows that appear in both `y` and `z`

```{r}
FILL_THIS_IN(y, z)
```

2. You collected the data in `y` on Day 1, and `z` in Day 2. 
   Make a data set to reflect that.

```{r}
FILL_THIS_IN(
  mutate(y, day = "Day 1"),
  mutate(z, day = "Day 2")
)
```

3. The rows contained in `z` are bad! Remove those rows from `y`.

```{r}
FILL_THIS_IN(FILL_THIS_IN, FILL_THIS_IN)
```
