# Filesystems and paths

Let's first load the built-in gapminder dataset and the tidyverse:

```{r}
library(gapminder)
library(tidyverse)
```

Next, let's filter the data to only 2007 and only the Asia continent and
save it to a new object.

```{r, eval = FALSE}
(gap_asia_2007 <- gapminder |> filter(year == 2007, continent == "Asia"))
```

We can write this to a comma-separated value (csv) file with the command:

```{r eval=FALSE}
write_csv(gap_asia_2007, "exported_file.csv")
```

But where did this file go? 
We should save the file in a sensible location. 
We need to practice controlling where R is running, where it looks for files, and where it writes out your files. 
To that end, let's review the working directory and RStudio Projects.


# The working directory and RStudio Projects

When you open R, it "runs" in some folder on your computer. 
This is the place it will look for files to import and write files as output.
Think about where your class files end up when you knit them.

If you have R/RStudio closed, and you open a .R or .Rmd file, R/RStudio will start in the folder holding that file.

If you open R/RStudio from the Windows Start menu, the Mac dock, the Mac Spotlight, etc., R/Studio will start in its default location (probably your user home directory, see Tools → Global Options → General → Default working directory...).

Write all of your R scripts assuming:

1.  You are running them by opening a fresh new R session

    -   Don't use `rm(list = ls())` to clean the workspace--the workspace is already clean
    -   You need to load required pacakges with `library()`
    -   Don't work on several different projects in one R session at the same time!

2.  You have all of the necessary packages installed

    -   Don't include `install.packages()` calls unless they are commented out or otherwise set not to run

3.  The script will run without human input

    -   You need to import or load any data you are working with
    -   Load data and write output using R commands, not `file.choose()`, `read.clipboard()`, the buttons in RStudio, etc.

4.  All of the needed files live in your project folder

    -   Write *relative* paths, rather than *absolute* paths
    -   `file.path("data", "cats-data_2020-03-04.xlsx")`
    -   Not:
        `/Users/brenton/Research/cats_project/data/cats-data_2020-03-04.xlsx`
        or
        `C:\\Users\\brenton\\Research\\cats_project\\data\\cats-data_2020-03-04.xlsx`\
        or`file.path("C:", "Users", "brenton", "Research", "cats_project", "data", "cats-data_2020-03-04.xlsx")`

5.  Your script might run on any system

    -   Write the paths to files using `file.path()` or `here::here()`, rather than typing out the whole path 
    -   This is because Windows and Mac/Linux have different syntaxes for file paths:

        -   Mac/Linux: `path/to/folder`
        -   Windows: `path\\to\\folder`

This approach has several advantages:

1.  Frictionless running on different computers
2.  Less breakage (e.g., if you move a folder around, the relative paths will still work)
3.  Less surprising or weird behaviors due to session crud
4.  Easy to tweak data/code and update results

# RStudio projects

RStudio projects can help with following these best practices. 

The best way to work with R is to have a separate Project for each project/paper you are working on. Switch RStudio to that Project whenever you are working on those files. 

You can you are working in a Project because its name will be in the RStudio title bar. 
One macOS, the Project name will also show on the Dock icon.


# here::here()

The *here* package is the best way to refer to files. 

The major function is `here::here()`. 
Like `file.path()`, `here::here()` lets you specify a path to a file and then adds the system-appropriate separators (`/` or `\\`). 

Where `here::here()` shines is that it figures out where the relative paths should start from. 
It looks around in the folders in your directory and finds the .Rproj file, then constructs the relative file paths from there.

The .Rproj file is a sign post that says "START HERE".

For example, create a new folder "data" in your participation folder. 
Then, save `gap_asia_2007` using the `here::here()` and `write_csv()` functions:

```{r, eval = FALSE}
write_csv(gap_asia_2007, 
          here::here("participation", "data", "gap_asia_2007.csv")
)
```

More details on `here` are available in this [short
article](http://jenrichmond.rbind.io/post/how-to-use-the-here-package/).
