# Reading data from disk

The same csv file that we just saved to disk can be imported into R again by specifying the path where it exists:

```{r, eval = FALSE}
dat <- read_csv(here::here("participation", "data", "gap_asia_2007.csv"))
dat
```

Notice that the output of the imported file is the same as the original tibble.
`read_csv()` was intelligent enough to detect the types of the columns. 
This won't always be true so it's worth checking! 
In particular, be on the lookout for any columns it imports as `col_character()`!

The [read_csv()](https://readr.tidyverse.org/reference/read_delim.html) function has many additional options including the ability to specify column types (e.g., is "1990" a year or a number?), skip columns, skip rows, rename columns on import, trim whitespace, and more.

To control the column types, use the `cols()` function:

```{r, eval = FALSE}
dat <- read_csv(
  here::here("participation", "data", "gap_asia_2007.csv"),
  col_types = cols(
    country = col_factor(),
    continent = col_factor(),
    year = col_date(format = "%Y"),
    .default = col_double() # all other columns as numeric (double)
  )
)
dat
```

By default, it leaves all columns as `col_guess()`, but it's better to be explicit.

Another important option to set is the `na` argument, which specifies what values to treat as `NA` on import. 
By default, `read_csv()` treats blank cells (i.e., `""`) and cells with `"NA"` as missing. 
You might need to change this (e.g., if missing values are entered as `-999`). 
Note that `readxl::read_excel()` by default only has `na = c("")` (no `"NA"`)!

```{r, eval = FALSE}
dat <- read_csv(
  here::here("participation", "data", "gap_asia_2007.csv"),
  col_types = cols(
    country = col_factor(),
    continent = col_factor(),
    year = col_date(format = "%Y"),
    .default = col_double() # all other columns as numeric (double)
  ),
  na = c("", "NA", -99, "No response")
)
dat
```


# Import a file from the web/cloud

## Import a CSV file from the internet

To import a CSV file directly from the web, assign the URL to a variable

```{r, eval = FALSE}
url <- "http://gattonweb.uky.edu/sheather/book/docs/datasets/magazines.csv"
```

and then apply read_csv file to the `url`.

```{r, eval = FALSE}
dat <- read_csv(url)
```

You can do this in one step if you like:

```{r, eval = FALSE}
read_csv("http://gattonweb.uky.edu/sheather/book/docs/datasets/magazines.csv")
```



## Import an Excel file (.xls or .xlsx) from the internet

First, we'll need the package to load in Excel files:

```{r, eval = FALSE}
library(readxl) 
```

Datafiles from this tutorial were obtained from:
<https://beanumber.github.io/sds192/lab-import.html#data_from_an_excel_file>

Unlike with a CSV file, to import an .xls or .xlsx file from the internet, you first need to download it locally.

*Note: The folder you want to save the file to _has to exist_!.*
*If it doesn't, _you will get an error._*

You can create the folder path in one of three ways:

  1. Create them directly in Finder/Windows Explorer
  2. Use the buttons in the Files tab in RStudio
  3. Use the `dir.create()` function:

```{r, eval = FALSE}
if ( !dir.exists( here::here("participation", "data") ) ) {
  dir.create( here::here("participation", "data"), recursive = TRUE )
}
```

Next, you download the file.
To download it, create a new object called `xls_url` and then use `download.file()` to download it to a specified path.

```{r, eval = FALSE}
xls_url <- "http://gattonweb.uky.edu/sheather/book/docs/datasets/GreatestGivers.xls"
download.file(
  xls_url, 
  here::here("participation", "data", "some_file.xls"), 
  mode = "wb"
)
```

**NOTE:** Don't assign the result of `download.file()`.

**NOTE:** The `mode = "wb"` argument at the end is really important if you are on Windows. 
If you omit, you will probably get a message about downloading a corrupt file.
More details about this behavior can be found [here](https://github.com/HenrikBengtsson/Wishlist-for-R/issues/30).


Naming a file "some_file" is extremely bad practice (hard to keep track of the files).
You should **always** give it a more descriptive name. 
It's also a good idea to avoid spaces in filenames.
You should come up with a system for naming your files and use it consistently. 
My file names look like this: `progdata_example-dataset_2021-03-02_bmw.csv`


Often, it's a good idea to name the file similarly (or the same as) the original file (sometimes that might not be a good idea if the original name is non-descriptive).

There's handy trick to extract the filename from the URL:

```{r, eval = FALSE}
file_name <- basename(xls_url)
download.file(
  xls_url, 
  here::here("participation", "data", file_name), 
  mode = "wb"
)
```

Now we can import the file:

```{r, eval = FALSE}
dat <- read_excel(
  here::here("participation", "data", file_name)
)
```


# Read in a sample SPSS file.

Let's load a sample SPSS file and work with it. 
Download the file from [here](https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/data/clevel.sav) and save it in your `data` folder in your class GitHub repo.

These data are a random subset of the data used in [this paper](https://doi.org/10.1016/j.jvb.2018.02.005). 
This was a study looking at personality traits that distinguish C-level executives from lower-level managers among men and women.

The subset of data here consists of 200 cases, with variables indicating: 

  1. The language of assessment (English, Dutch, or French) 
  2. Gender 
  3. C-level or not 
  4. Extraversion level, as well as 4 facet traits (Leading, Communion, Persuasive, Motivating)

Let's load in the data using the `haven` package.

```{r, eval = FALSE}
(clevel <- haven::read_spss(
  here::here("participation", "data", "clevel.sav")
)
)
```

Notice that this tibble looks a little different for the `language` and `gender` variables than normal. 
It has **labels** for the numeric values. 
This format is what SPSS uses, but it's not standard for R.
Let's convert those variables, and `isClevel` as well, to factors:

```{r, eval = FALSE}
clevel_cleaned <-
  clevel |> 
  mutate(language = as_factor(language),
         gender = as_factor(gender),
         isClevel = factor(isClevel, 
                           levels = c(0, 1), 
                           labels = c("No", "Yes"))
  )
print(clevel_cleaned)
```

Notice how the variables are now factors with labels as the entries, instead of the original code numbers.
