# Saving files

## Saving data frames

Let's save this file as a CSV file so that it's a smaller file and easier to import again in the future.

```{r, eval = FALSE}
write_csv(
  clevel_cleaned, 
  here::here("participation", "data", "clevel_cleaned.csv")
)
```

## Saving plots

Now let's make a plot.

```{r, eval = FALSE}
clevel_plot <-
  clevel_cleaned |> 
  mutate(isClevel = recode(isClevel, 
                           No = "Below C-level", 
                           Yes = "C-level"),
         gender = recode(gender,
                         Female = "Women",
                         Male = "Men")) |> 
  ggplot(aes(paste(isClevel, gender, sep = "\n"), Extraversion, color = gender)) +
  geom_boxplot() +
  geom_jitter(height = .2) +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  ggtitle("Extraversion Stan Scores") +
  scale_y_continuous(breaks = 1:9) +
  ggthemes::theme_fivethirtyeight()

clevel_plot
```

Let's save the plot in several formats. 
This is useful if we want to use the plot outside of Markdown. 
Save plots using the `ggsave()` funnction.

`ggsave()` has many options. 
See the help function `?ggsave` for full details. 
The main arguments are:

  - `filename`
  - `plot`
  - `width` and `height` (inches by default)
  - `dpi` (dots per inch; for bitmap formats).

`ggsave()` will try to guess what format you want based on the file name. 
If you want, you can specify a specific format or R graphics device to save with using the `device` argument.

```{r, eval = FALSE}
dir.create(here::here("participation", "output", "figures"), recursive = TRUE)
ggsave(here::here("participation", "output", "figures", "clevel_extraversion.svg"), 
       clevel_plot,
       height = 6,
       width = 6
)
```

You can save to several formats. 
Generally, work with a *vector* format like `.svg`, `.eps`, or `.pdf`. 

Vector graphics represent the image as a series of data points and equations. 
This means that they can be made smaller or larger or zoomed in on without damaging the image quality.

If you can't use a vector format for some reason, you can also export to a *bitmap* format. 
Bitmap graphs represent the image as colored dots or pixels. 
This means that the image quality will suffer if you make the image larger or zoom in on it (making it smaller can also sometimes compromise quality). 
With bitmap images, you need to be concerned with *resolution* (how many pixels/dots per inch when printed). 
Always use at least **300 DPI** resolution.

There are several popular bitmap image formats. 

  - `.tiff`/`.tif` 
    - Highest quality, but also the largest. 
    - Use it for print graphics, but you should probably avoid it for images to be hosted on the web. 
  - `.png` 
    - A bit smaller, and it should be your go to for charts, figures, line drawings, etc. 
    - More complex images (e.g., photos) can get pretty big with `.png`, though. 
  - `.jpg`/`.jpeg`
    - Probably the most popular bitmap format.
    - Works well for photographs hosted on the web, but its compression often makes line drawing and charts look terrible.
    - `.jpg`/`.jpeg` also degrades in quality each time you edit/save it, so don't use it for images you intend to edit. 
  - `.gif`
    - Generally avoid `.gif` unless you are making an animation or you need very small file size for a simple image. 
    - `.gif` supports very few colors, so always check your image quality after making a `.gif`.

Let's save to some other formats:

```{r, eval = FALSE}
ggsave(here::here("participation", "output", "figures", "clevel_extraversion.pdf"), clevel_plot, height = 6, width = 4)
ggsave(here::here("participation", "output", "figures", "clevel_extraversion.tiff"), clevel_plot)
ggsave(here::here("participation", "output", "figures", "clevel_extraversion.png"), clevel_plot)
```
