# Entering Data Manually

## Making data.frames or tibbles

In base R, we can make data frames using the `data.frame()` function. 
The tidyverse version is `tibble::tibble()`. 
`tibble()` is stricter by not allowing recycling unless the vector is of length 1.

Good:

```{r}
dat <- tibble(
  x = 1:6,
  y = min(x)
)
```

Bad:

```{r, error = TRUE}
dat <- tibble(
  x = 1:6,
  y = 1:2
)
```


Manual construction of tibbles is convenient with `tibble::tribble()`:

```{r}
dat <- tribble(
  ~Day, ~Breakfast,
  1, "Apple",
  2, "Yogurt",
  3, "Yogurt"
)
```

## datapasta

The *datapasta* package helps to reproducibly copy-paste data from spreadsheets into R. 
`datapasta` uses `tribble()`.

```{r, eval = FALSE}
install.packages("datapasta")
```

After you install the package, you can set keyboard shortcuts in RStudio for its functions:

  1. Click Tools → Modify Keyboard Shortcuts…
  2. Search for 'paste'.
  3. Set keyboard shortcuts for "Paste as tribble", "Paste as vector", and other functions if you like.

Now, try out pasting a vector and a tibble.
