```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
```

# Customizing plots!

You can customize all sorts of stuff in your ggplot plots!

1.  Scales
  -   x and y ✔
  -   Colors ✔
  -   Shapes
  -   Line types
  -   Alpha, Size

2.  Guides
  -   Legend (legend, bins, colorbar, colorsteps)
  -   Axis
  -   None

3.  Text
  -   Axis labels
  -   Geom labels
  -   Annotations
  -   Titles and captions

4.  Themes (design features)
  -   Background
  -   Grid lines
  -   Text styling
  -   Alignment

# Scales

## Shape scales

In addition to distinguishing groups or values by color, you can also distinguish groups by **shape**.
You can see a list of R's built-in plot shapes [here](http://www.cookbook-r.com/Graphs/Shapes_and_line_types/).

You can use `scale_shape_*()` with **discrete** and **binned** variables, but not *continuous* variables.

R has a set of built-in plot shapes:

  -   0--14: Empty shapes --- outline `color`, no `fill`
  -   15--20: Solid shapes --- solid `color`, no separate `fill`
    -   Default ggplot shape is 19
  -   21--25: Filled shapes --- outline `color`, separate `fill`

```{r}
ggplot(mtcars) +
  aes(x = hp, y = mpg, shape = factor(cyl)) +
  geom_point(size = 4, color = "blue", fill = "yellow") +
  scale_shape_manual(values = c(1, 19, 25)) +
  theme_minimal()
```

You can also give a character vector to plot those characters. 
Each element must be a single character long.

```{r}
ggplot(mtcars) +
  aes(x = hp, y = mpg, shape = factor(cyl)) +
  geom_point(size = 4, color = "blue", fill = "yellow") +
  scale_shape_manual(values = c("4", "6", "8")) +
  theme_minimal()

ggplot(mtcars) +
  aes(x = hp, y = mpg, shape = factor(cyl)) +
  geom_point(size = 4, color = "blue", fill = "yellow") +
  scale_shape_manual(values = c("😸", "😾", "🙀")) +
  theme_minimal()
```

`scale_*_identity()` is a special `scale` that uses the value of the variable directly on the plot.
With shape, you can use it to plot the values if they are single character. 
(Use one of the text geoms if you want multi-character shapes.)

```{r}
ggplot(mtcars) +
  aes(x = hp, y = mpg, shape = factor(cyl)) +
  geom_point(size = 4, color = "blue", fill = "yellow") +
  scale_shape_identity() +
  theme_minimal()
```

You can also use it with `color` or `fill` to store the color codes directly in the data.

```{r}
tibble(
  scale = forcats::as_factor(c("ES", "A", "C", "Ex", "O")),
  score = c(5, 8, 4, 6, 3),
  .fill = c("#CD0BBC", "#DF536B", "#2297E6", "#61D04F", "#F5C710")
) %>% 
  ggplot() +
  aes(x = scale, y = score, fill = .fill) +
  geom_path(group = "n") + 
  geom_errorbar(aes(ymin = score - .5, ymax = score + .5), width = .05) +
  geom_point(shape = 21, size = 5) +
  scale_fill_identity() +
  ylim(0, 10) + 
  theme_minimal()
```

## Linetype scales

Line type is like shape. You can specify line type for **discrete** and **binned** variables, but not *continuous* variables.

You can see a list of R's built-in line types [here](http://www.cookbook-r.com/Graphs/Shapes_and_line_types/). 
You can refer to line types by number or name.

R's built-in line types:

| Number | Name       |
|--------|------------|
| 0      | 'blank'    |
| 1      | 'solid'    |
| 2      | 'dashed'   |
| 3      | 'dotted'   |
| 4      | 'dotdash'  |
| 5      | 'longdash' |
| 6      | 'twodash'  |

## Alpha and size scales

`scale_alpha_*()` (transparency) and `scale_size_*()` work similarly to `scale_color_*()`.
You may want to set the `range` argument to control the minimum and maximum alpha/size used.
Alpha can range between 0 and 1.


# Guides

Guides are the instructions for how to read the scales on your plot.
The two major kinds of guides are **axes** and **legends**.
You can customize the appearance of your guides using the `guides()` and `guide_*()` functions.

```{r}
p <- data.frame(
  x = 1:5, y = 1:5, p = 1:5, q = 1:5, r = factor(1:5)
  ) %>% 
  ggplot() +
  aes(x, y, colour = p, size = q, shape = r) +
  geom_point() +
  scale_size_continuous(range = c(4, 8)) +
  theme_classic()
  
p

# Legends will be combined if they have the same type, values, and titles 
p + 
  guides(
    colour = guide_legend(title = "title"),
    size = guide_legend(title = "title"),
    shape = guide_legend(title = "title")
 )

# If all you want to change is the title of the guides, you can use labs()
p + 
  labs(
    color = "title",
    size = "title",
    shape = "title"
  )
```

## Legend guide options

There are four styles of legend guides available:

1. `guide_legend()`
  - Basic legend guide.
  - Use this one most of the time.
2. `guide_bins()`
  - Use if you've got binned scales.
3. `guide_colorbar()`
  - Nicely show a continuous color gradient.
4. `guide_colorsteps()`
  - Use with binned color/fill scales.
  - Can also be useful if you have continous scales, but want to pick out useful reference levels.
  
```{r}
p1 <- p + 
  guides(
    colour = guide_legend(title = "legend"),
    size = "none",
    shape = "none"
  )
p2 <- p + 
  guides(
    colour = guide_bins(title = "bins"),
    size = "none",
    shape = "none"
  )
p3 <- p + 
  guides(
    colour = guide_colorbar(title = "colorbar"),
    size = "none",
    shape = "none"
  )
p4 <- p + 
  guides(
    colour = guide_colorsteps(title = "colorsteps"),
    size = "none",
    shape = "none"
  )

p1 + p2 + p3 + p4
```

### Legend arguments

1. `title`
  - Set the title of the legend guide
  - Also impacts legend merging
2. `direction` 
  - `"vertical"` or `"horizontal"`
3. `nrow` and `ncol`
  - Control number of rows and columns of legend
4. `reverse`
  - `TRUE` or `FALSE`: Should the order of the legend values be reversed?
5. `order`
  - If multiple legend guides are shown separately, what order should they appear in?
  - A number between 1 and 99

To place the legend somewhere other than the right side, add `theme(legend.position = "bottom")` to the plot:

```{r}
p + 
  guides(
    colour = guide_legend(title = "title"),
    size = guide_legend(title = "title"),
    shape = guide_legend(title = "title")
  ) +
  theme(legend.position = "bottom")
```

## Axis guides

`guide_axis()` controls how the axes appear. 
The main use is to control label overlap and change the axis position.
Some useful arguments are:

1. `title`
  - Set the title of the axis guide
2. `check.overlap` 
  - Drop some labels if they overlap?
  - Mostly usful if the values are ordered.
3. `angle`
  - Rotation in degrees of axis labels
  - This is usually very hard to read and should be avoided.
4. `n.dodge`
  - Put labels on multiple rows/columns to avoid overlap?
  - A number indicating how many rows/columns
5. `position`
  - Where to put the axis?
  - `"top"`, `"bottom"`, `"left"`, `"right"`
  

## Hide guides

You can hide a guide with `guide_none()`.

```{r}
p + 
  guides(
    colour = guide_legend(title = "Legend Title"),
    size = guide_none(),
    shape = "none"
  ) +
  theme(legend.position = "bottom")
```

