# All the plot things

```{r}
library(ggplot2)
library(dplyr)
library(patchwork)
```

## Basic elements of ggplot

1. Data
2. Aesthetics (mapping)
3. Geoms
4. Facets
5. Stats

## Changing the `stat`

In the process of connecting data -> mapping, there is some **statistical computation**. 
In many cases, this computation is `identity()`. For example, with `geom_point()`, it plots the points directly based on `x` and `y` values.
Other times, there is more of a computation.

  - `stat_count()`
  - `stat_bin()`
  - `stat_density()`
  
`geom_*()` functions usually come with a reasonable default `stat_*()` and subsequent mapping.
For example, `geom_histogram()` does `stat_bin()` to bin the continuous values, then `aes()` maps `y` to the count in each bin.

But sometimes you might want to change this mapping after the stat.
Do that with `after_stat()` inside `aes()`.
  
```{r}
bw <- .3

h1 <- ggplot(mpg) +
  aes(x = displ) +
  geom_histogram(binwidth = bw, fill = "skyblue")

## histogram with density
h2 <- ggplot(mpg) +
  aes(x = displ) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = bw, fill = "skyblue")

## histogram with proportion
h3 <- ggplot(mpg) +
  aes(x = displ) +
  geom_histogram(aes(y = after_stat(count / sum(count))),
                 binwidth = bw, fill = "skyblue")

h1 + h2 + h3

## histogram with density curve
ggplot(mpg) +
  aes(displ) +
	geom_histogram(aes(y = after_stat(count)), 
	               binwidth = bw, fill = "skyblue") +
	geom_density(aes(y = after_stat(count) * bw)) +
	theme_minimal()
```


## Customizing plots!

You can customize all sorts of stuff in your ggplot plots!

1. Scales
  - x and y
  - Colors
  - Shapes
  - Line types
  - Alpha
2. Guides
  - Legend (legend, bins, colorbar, colorsteps)
  - Axis
  - None
3. Text
  - Axis labels
  - Geom labels
  - Annotations
  - Titles and captions
4. Themes (design features)
  - Background
  - Grid lines
  - Text styling
  - Alignment

### Scales

A ggplot is constructed with a flow like this:

> data -> stat -> scale -> mapping -> render

The "scale" step concerns how the computed data/stat corresponds to numbers/values on the plot (e.g., "20 mpg" corresponds to a point "3 units up on the y axis"; "4 cylinder" corresponds to "blue").

You control how values are scaled using the `scale_*()` functions.
Almost every aesthetic you can map to data (e.g., `x`, `y`, `color`, `fill`, `alpha`, `linetype`, `shape`, `size`, `radius`) has corresponding scale functions.

These functions all have the following form:

> `scale_[aesthetic]_[type]`

The `scale_*()` functions include many arguments to control various features of the scale, depending on the aesthetic.

#### Position scales

The scales for `x` and `y` control positioning, limits, breaks, labels, titles, etc. of the axes. 

  - `scale_x_continuous()`
  - `scale_x_discrete()`
  - `scale_x_binned()`

##### Arguments

1. `name()`
  - The name of the scale. Used for the axis/legend title.
  - Give multiple scales the same name to combine them.
  - `scale_x_continuous(name = "MPG")`
2. `breaks`
  - Where to break the scale (e.g., where to bin or where to place labeled tick marks)
  - To override the default computed breaks, pass a vector of values
  - (There are more advanced options as well)
  - `scale_x_continuous(name = "MPG", breaks = c(0, 10, 20, 30))`
3. `labels`
  - Override the existing axis labels
  - Usually given as a function that transforms the text
  - e.g.,
    - Numbers and Dates
      - scales::label_percent
      - scales::label_dollar
      - scales::label_date
      - scales::label_comma
      - scales::label_pvalue
    - Text
      - stringr::str_to_upper
      - stringr::str_to_sentence
      - scales::label_parse / scales::label_math
4. `limits`
  - The minimum and maximum values for the scale
  - e.g., `scale_x_continuous(limits = c(40, 100))`
5. `oob`
  - What to do with values outside the `limits`
  - Defaults
    - Continuous: `scales::oob_censor`
    - Binned: `scales::oob_squish`
  - `scales::oob_keep`: Like zooming, don't remove or change values.

```{r}
ggplot(mtcars) + 
  aes(x = hp, y = mpg, color = factor(cyl), shape = hp > 200) + geom_point(size = 3) + scale_x_continuous(limits = c(0, 200), oob = scales::oob_squish)
```


6. `trans`
  - A transformation function to apply to each value of the scales
  - Must be a quoted function name
  - e.g., "sqrt" or "rev" or "log10"
  - `scale_x_continuous(trans = "sqrt")`
  - `scale_x_sqrt()`, `scale_x_reverse()`, and `scale_x_log10()` are shortcuts
7. `position`
  - "top", "bottom", "left", or "right"
  - Can also be set in `guide_axis()`
8. `guide`
  - A function controlling formatting of the guide
  - I usually set in `guides()` instead
9. `na.translate`
  - TRUE or FALSE, for discrete scales, plot NA?
  - `scale_x_discrete(na.translate = FALSE)`
10. `na.value`
  - What value to replace `NA` with.
  - `scale_x_discrete(na.translate = TRUE, na.value = "(Missing)")`

#### Color scales

There are a variety of ways to specify colors in R, as well as several nice built-in color palletes.

Some resources on color tools in R:

  1. [R color cheatsheet](https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf)
  2. [colorspace package](https://colorspace.r-forge.r-project.org/articles/colorspace.html)
  3. [R cookbook colors](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/)
  4. [Another nice ggplot color guide](https://www.datanovia.com/en/blog/ggplot-colors-best-tricks-you-will-love/)
  5. [Colormind](http://colormind.io/)
  6. [coolors](https://coolors.co/)
  7. [Color Designer](https://colordesigner.io/)
  
Reminder:

  - `color`: The outline color for lines/points/regions
  - `fill`: The fill color for regions and point-shapes with a filling (shapes 21-25)
  
The color scale functions you should use in ggplot all have the following form:

> `scale_[aesthetic]_[type + pallete]`

For example:

  **Custom Gradients, Color Wheels, and Palettes**
  - Continuous: `scale_color_gradient()`
  - Continuous: `scale_color_gradient2()`
  - Continuous: `scale_color_gradientn()`
  
  - Binned: `scale_color_steps()`
  - Binned: `scale_color_steps2()`
  - Binned: `scale_color_stepsn()`
  
  - Discrete: `scale_color_hue()`
  - Discrete: `scale_color_manual()`
  
  **Viridis**
  - Continuous: `scale_color_viridis_c()`
  - Binned: `scale_color_viridis_b()`
  - Discrete: `scale_color_viridis_d()`
  
  **Color Brewer**
  - Discrete: `scale_color_brewer()`
  - Continuous: `scale_color_distiller()`
  - Binned: `scale_color_fermenter()`
  
  **Greyscale**
  - `scale_color_grey()`

  **ggthemes**
  - `ggthemes::scale_color_excel()`
  - `ggthemes::economist()`
  - `ggthemes::scale_color_fivethirtyeight()`
  - `ggthemes::scale_color_colorblind()`
  
  **Other packages**
  - _wesanderson_
  - _ggsci_

##### Arguments

All of the arguments that can be used with position scales can also be used with color scales.


1. `name()`
2. `breaks`
3. `labels`
4. `limits`
5. `oob`
6. `trans`
7. `position`
8. `guide`
9. `na.translate`
10. `na.value`

**Color-specific arguments**

Color scale functions also have arguments to control how the palettes are constructed.
The exact arguments vary by scale function.
Let's look at a few.

`scale_color_viridis_*()`

1. `begin` and `end`
  - Values between 0 and 1 saying where on the scale to start and stop.
2. `direction`
  - Which end to start from?
  - 1 or -1
3. `option`
  - Which of the viridis family palettes to use?
  - "A", "B", "C", "D", "E"
4. `aesthetics`
  - Which aesthetics to apply the palette to?
  - Useful to do `aesthetics = c("color", "fill")`

  
`scale_color_brewer()`

1. `direction`
  - Which end to start from?
  - 1 or -1
3. `type`
  - Which kind of color scale to use?
  - "seq" (sequential), "div" (diverging), "qual" (qualitative)
4. `palette`
  - Which specific palette to use?
  - Can be a number or one of the named palettes (see `?scale_color_brewer()`)
5. `aesthetics`


`scale_color_manual()`

For specifying your own discrete palettes.

1. `values`
  - A character vector of colors names or hex codes to use.
  - If the vector has names, it will match data values to names.
  - Must be at least as long as the number of unique values.
2. `aesthetics`

```{r}
p <- ggplot(mtcars, aes(mpg, wt)) +
  geom_point(aes(colour = factor(cyl)), size = 8)
p + scale_colour_manual(values = c("4" = "red", "6" = "blue", "8" = "darkgreen"))

pal <- palette.colors("Okabe-Ito", n = 10)
p + scale_colour_manual(values = pal)

pal <- palette.colors("R4", n = 4)
p + scale_colour_manual(values = pal)
```

`scale_color_gradient()`

For specifying your own gradient.

```{r}
scale_colour_gradient(
  ...,
  low = "#132B43",
  high = "#56B1F7",
  na.value = "grey50",
  aesthetics = "colour"
)

scale_colour_gradient2(
  ...,
  low = muted("red"),
  mid = "white",
  high = muted("blue"),
  midpoint = 0,
  na.value = "grey50",
  aesthetics = "colour"
)

scale_colour_gradientn(
  ...,
  colours,
  values = NULL,
  na.value = "grey50",
  aesthetics = "colour",
  colors
)
```


`scale_color_identity()`

A special scale.
It doesn't map colors to a specific value, but uses the variable values directly.
Useful if you want to store the color codes directly in the data.
