# Introduction to GitHub and Version Control

__Announcements__:

- The first assignment is due for peer review on Monday, final on Wednesday
- The first peer review is due on Tuesday


## Outline

We'll cover four topics today:

- Git and GitHub
- Version control
- Submitting assignments
- Asking questions


## Learning Objectives

By the end of today's class, you will be able to:

- Understand the benefits of a version control workflow 
- Commit file changes to GitHub
- Open and respond to pull requests on GitHub
- Navigate the commit history of a repository and a file on GitHub


## Resources

If you want to learn more about today's topics, check out:

- The [GitHub guide](https://guides.github.com/) has lots of info about GitHub.
  If you do go here, I recommend you start with "Hello, World!"](https://guides.github.com/activities/hello-world/). 
- Dr. Jenny Bryan's ["How to get unstuck"](https://progdata.netlify.com/help-general/) page is useful for getting help online.


## Making a GitHub Repository

We will be using [GitHub](https://github.com) a lot in this course.

GitHub is similar to Dropbox, OneDrive, or other cloud storage services.
There are 3 key benefits of GitHub compared to those other services.

  1. GitHub is a _version control_ platform. 
     It is designed to **track** exactly **what changes you make** and **when you make them**.
     You can browse the history 
  2. GitHub is designed for code.
     You can track what changes you make **line by line**
     You can integrate your code with automated tools for error checking, testing, etc.
  3. GitHub is designed for _collaborative coding_.
     
       - One master copy
       - People can make simultaneous edits
       - No emailing files back and forth
       - Tools for checking and verifying each other's code
       
     We will practice using GitHub for collaboration through the weekly homework.
     
Let's make a GitHub repository that you will use for this class.
Follow this link: <https://classroom.github.com/a/bUKffmtV>

This will create a repository for you or take you to your existing repository.

     
## Navigating a GitHub Repository

Let's practice some vocabulary.

  1. You each have a GitHub **account**.
  1. You each have made a homework **repository** or **repo**.
  
     - A repository can be *public* or *private*.
     - You can have many repositories for your account.
     - It's a good idea to **one repository for each project**.
     - (Sign up for the GitHub Student Pack to get free unlimited private repositories.)
     
  1. A repo can contain many files and folers (or "directories").
  1. When you make a change to a file in your repo, this is called **committing** a change (or making a **commit**).
  
Let's take a look at various parts of a GitHub repo.

  1. Files/folders ("code")
  1. History
  1. Issues
  1. Branches
  1. Pull requests

Let's practice editing a file with GitHub.
Make a new file on your participation repository and edit it:

    - Click on the "Create New File" button on your repository's home page.
    - Call it `navigating_github.md`
    - Write something (e.g., what's your favorite color?). 
    - Commit ("save") the file by clicking on green "commit new file" button at the bottom of the page.
    - Now type something else (e.g., what's your favorite animal?) and commit the change.
    
Now look at your commit history in the "History" tab.
  
You can download a copy of your repository to your computer (**cloning** it), make changes (commits) there, then **push** them back to GitHub.
You do this with a **git client** application, such as the one provided by GitHub: [GitHub for Desktop](https://desktop.github.com)

After GitHub for Desktop is installed, you can clone the repo to your computer using the big green "Code" button on the GitHub page.
You can also clone a repository and work with it directly in RStudio.
Let's do that now.


## RStudio Projects

In RStudio, click File → New Project… → Version Control → Git. In the popup:

  1. Enter the URL for your GitHub repository
  1. Choose where you want the folder to be stored on your computer
  1. Give the folder (directory) an understandable name
  
This will do two things:
  1. Download ("clone") the GitHub repo to your computer and start tracking it
  1. Create an RStudio Project file

### The working directory

When you open R, it "runs" in some folder on your computer. 
This is the place it will look for files to import and write files as output. 
Think about where your Rmd output files end up when you knit them.

If you have R/RStudio closed, and you open a .R or .Rmd file, R/RStudio will start in the folder holding that file.

If you open R/RStudio from the Windows Start menu, the Mac dock, the Mac Spotlight, etc., R/Studio will start in its default location (probably your user home directory, see Tools → Global Options → General → Default working directory…).

When I say "R/Studio will start in…", what I am referring to is R's "working directory". 
Like I say above, this is the place R will look for files to import and write files as output. 
You can check what R's current working directory is using the `getwd()` function:

```{r}
getwd()
```

You can also change the working directory using the `setwd()` function:

```{r, eval = FALSE}
setwd(file.path("path", "to", "folder"))
```

Do not use `setwd()`! 
You should always write your R scripts so that the entire project is self-contained in a folder. 
All of the scripts, folders, data, output, etc. should all "live" within this project folder. 

We will talk a lot about how to do this throughout the semester.
For now, we will start by working with RStudio Projects to make this easier.

When you double click on a .Rproj file, it:

  1. Opens a new fresh R session, with
  2. The working directory set to the location of the .Rproj file, and
  3. No connection whatsoever to any other R sessions you already have open
  
You can also set specfic options for each RStudio project (e.g., number of spaces to insert when you type Tab, etc.).

Let's practice closing RStudio and re-opening it by opening the RStudio project.
Run `getwd()` to see where R is running. 


## The Version Control Workflow

Now, let's practice the workflow to work with your files with GitHub.

### Editing a file and making a commit

Open the `README.md` file in your local git folder by clicking on it in the Files pane. 
`README.md` is a special file that will show when viewing a folder on GitHub.
Use README files to describe the contents of a folder and what's going on there.

Type "Hello world" and save the file.

Now, open the Git tab on the right hand side of the RStudio window. 

This pane shows you the files in the git fold that **have changed since the last commit**. 
Any files shown here have had changes made.
(Remember, you need to save the file in RStudio [so the title on the Source tab isn't blue] before they appear in this list.)
Changes shown here are not saved in the repo history until you **commit** them.

  1. The "Path" column gives the filename (and folder path)
  1. The "Staged" column indicates whether the changes will be saved in the next commit.

     - Check the box to "stage" the file (add it to the commit)
     - Uncheck the box to "unstage" it (remove it from the commit)
     
  1. The "Status" column indicates how a file has changed
  
     - "M" for modified
     - "A" for added
     - "R" for removed
     
You can view the changes in more detail by clicking the "Diff" button.
Here you can view individual line changes, stage/unstage files, and view the commit history.

If you want to undo any changes and revert back to the previous committed version of a file, select it in the Diff list and click the "Revert" button.

To **commit** your changes, type a **commit message** describing what you've changed in the message box on the right.
Always give informative commit messages (help out future you!).
When you are ready, click the "Commit" button.

Now, you've made the commit locally, but you need to "push" it to GitHub so that it shows up there as well. 
Click the "Push" (up arrow) button.

Now go check out the file on GitHub online.

### Fetching or Pulling Changes from the Remote Repo

One of the amazing things about git is that it can track changes made to a file at different locations or on different computers and reconcile them together.
(This is how it is so useful for collaboration!)

Let's see how that works.

  1. First, let's make a change directly on the GitHub website.
     Edit `README.md` and type "oops!" at the top.
     Commit your change.
  1. Go back to RStudio.
  1. Click the "Pull" (down arrow) button in the Git tab.
  1. See how the file on your local computer changed to reflect the change you made online.
  
Let's fix that "oops!".
Delete it, commit your change (give an informative commit message), and push your changes back online.

### The General Workflow

You will use this basic workflow througout the semester (and your coding career).

When you sit down to start working:

  1. Open you RStudio Project.
  1. Click the Pull button to fetch remote changes and get your local repo copy up to date.
  1. Make changes.
  1. Commit your changes.
  1. Push your commits back up to the remote repo (GitHub).
  
You should get in the habit of **commit early, commit often**.
Don't wait until you are completely done with your work to commit the changes.
Make small commits as you go.
This makes it much easier to go back if you accidentally break something and need to revert.

Next week, we will practice what to do if something goes wrong with this workflow and you need to fix things (e.g., merge conflicts, reverting commits).


## Branches and Pull Requests

Sometimes, it's a good idea to work on changes to one part of your code separately before putting them back into your master document. 
For example, let's say youwant to experiment with a new analysis without messing up your existing work.

To do that, we use **branches**. 
A branch is a copy of the repo that sits separately from your `master` branch. 
Git keeps track of how each of your branches differs.

  - When did a branch break off from `master`?
  - What commits have been made to the branch since the break?
  - What commits have been made to `master` since the break?
  
Let's take a look at some branches on the [psychmeta repo](https://github.com/psychmeta/psychmeta).

After you make changes to the branch, at some point, you probably want to **merge** the branch back into `master`. 
On GitHub, we do this using a **pull request**.

  - That name means you want the parent branch (`master`) to _pull_ in all the commits made on child branch. 
    You _request_ that this happens from the  owners or collaborators on the repo (for your homework, this is yourself).
  - Once a pull request has been opened, it's a good time to review the changes to the code, run tests, make comments, and correct any errors. 
    You can review yourself and also request code reviews from others by username.
  - You can make additional changes to the pull request by pushing more commits to the child branch.
  - When you are ready, **merge** the pull request into the parent branch.
  - [Example from psychmeta](https://github.com/psychmeta/psychmeta/pull/75).
  - For more info on pull requests, see this [GitHub tutorial](https://help.github.com/articles/about-pull-requests/).

There are many reasons you may want to branch. Here are some:

- A collaborator wants to make a change to the repo, but the end product of the change requires review from collaborators.
- You want to make changes, but don't want to "deploy" the changes until later (such as if pushing to GitHub triggers a website build).
- If you want to try something "risky", it's just safer to work on a branch.

### Making a Branch

Let's organize your homework repo in a branch.

  1. Create a new branch locally by clicking the "Branch" button in RStudio (next to the "master" branch name)
  
     - Call this branch "organizing"

  1. Open your repo folder in your computer's file system.
  
  1. Make a folder called "participation" and move all of the files besides `README.md` into it.
  1. Make a folder called "hw01". 
     This will be where you save your files for your first homework.
  1. Stage and commit your changes.
  
Now let's see how the branches work.
Switch back and forth between the `master` and `organizing` branches.
See how the organization changes happened on `organizing` but not on `master`.

### Making a Pull Request

Now, let's merge the changes from `organizing` back to `master`.

  1. Go to your repo on GitHub.
  
     - **Pro tip:** I link to add a shortcut link to the GitHub page for the repo to the folder to make it easier to get to GitHub from RStudio.
     
  1. Switch to the `organizing` branch.
  1. Click "Pull Requests" and then the green "New pull request" button.
  1. Select the approprite "base" (`master`) and "compare" (`organizing`) branches.
  1. Type a description of the changes made in the box.
  1. Click "Create pull request".
  
Once the pull request is open, your changes are ready to be reviewed, then merged back into `master`.

For your homework, you will make a new branch called `hw01`. 
Do your work on this new branch.
When you are ready for peer review, open a pull request and tag your assigned reviewers by clicking the "Request review" link and typing their names.

### Commenting on a Pull Request

Once a pull request is open, folks can review your code and make comments.
You will review two of your classmates' code each week.
You can find a peer review template on the course website.

Be sure your comments are productive and constructive.
Be kind!

You can mention specific people (e.g., me) by typing `@` and then their username.

You can make general comments and feedback, as well as comments on specific lines (and even suggest specific edits).

### Revisining and Merge a Pull Request

After you receive comments, make any necessary changes to your code.
You can update a pull request by making more commits to the same branch (e.g., to the `hw01` branch).
No need to make a new branch here.

Once you have finished updating, **merge** the pull request by going to it on GitHub online, and clicking the "Merge" button.

After you have merged the pull request, it's a good idea to delete the branch.


## This week's homework

### Assignment

The assignment is posted on the [Assignments page](https://progdata.netlify.app/evaluation/assignments/) of the course website.

Your tasks are to create 2 products:

  1. A `README.md` file for your homework repository to introduce yourself and show off some Markdown formatting skills.
     Place this in your repo's root (home/base) folder.
  2. An RMarkdown document exploring a dataset, such as `palmerpenguins::penguins`, `gapminder::gapminder`, or `psych::bfi`.
     Place this in the `hw01` folder.


For this assignment: 

  - Follow the instructions in the HW01 description carefully. 
  - Commit your code to your homework repo on GitHub **on a new branch called `hw01`**. 
  - When you are ready for your peer review (by Monday at 23:59!), create a  **Pull Request** to prepare to merge this branch back into your `master` branch.
  - Request code reviews from the 2 people assigned to you (see the [Peer Review page](https://progdata.netlify.app/evaluation/hw01/pr01/)).
  - Submit your peer review comments to your classmate by Tuesday at 23:59.
    - After you make your PR comments, copy a link to the PR page and submit this through Teams.
  - Merge your homework branch back into `master` by Wednesday at 23:59.
    - Afer you merge your PR, copy a link to the `hw01` folder and submit this through Teams.
    
If you run into any issues getting R or GitHub to work, please ask me!
Don't spend hours bashing your head against your keyboard!
