# Tidy Data and Pivoting


```{r, warning = FALSE, message = FALSE}
library(tidyverse)
```


## Orientation


### Worksheet

You can find a worksheet template for today
[here](https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/tutorials/s06_pivot-exercise.Rmd).


### Today

Today's concept is __tidy data__ and the `tidyr` package.

- Reshaping data by pivoting with `tidyr::pivot_longer()` and `tidyr::pivot_wider()`.
- Making tibbles using `tibble::tibble()` and `tidyr::expand_grid()`.


### Resources

For concepts of tidy data: 

- [Jenny Bryan's intro to tidy data](https://github.com/jennybc/lotr-tidy/blob/master/01-intro.md) is short and sweet.
  - the repo this links to has some useful exercises too, but uses the older `spread()` and `gather()` functions.
- `tidyr` [vignette on tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html).
- [Hadley's paper on tidy data](https://vita.had.co.nz/papers/tidy-data.pdf) provides a thorough investigation.

For pivoting with `tidyr`, check out the [pivot vignette](https://tidyr.tidyverse.org/articles/pivot.html).

I also recommend reading the new additions that come with the new `tidyr` Version 1.0.0 in [this tidyverse article](https://www.tidyverse.org/articles/2019/09/tidyr-1-0-0/).


## Tidy Data

A data set is _tidy_ if:

- Each row is an __observation__;
- Each column is a __variable__;
- Each cell is a __value__.

This means that each value belongs to exactly one variable and one observation.

Why bother? Because doing computations with untidy data can be a nightmare. Computations become simple with tidy data. 

Whether or not a data set is "tidy" depends on the type of analysis you are doing or plot you are making. 
It depends on how you define your "observation" and "variables" for the current analysis.

```{r, echo = FALSE}
haireye <- as_tibble(HairEyeColor) %>% 
  count(Hair, Eye, wt = n) %>% 
  rename(hair = Hair, eye = Eye)
```

As an example, consider this example derived from the `datasets::HairEyeColor` dataset, containing the number of people having a certain hair and eye color.

If one observation is identified by a _hair-eye color combination_, then the tidy dataset is:

```{r}
haireye %>% 
  DT::datatable(rownames = FALSE)
```

If one observation is identified by a _single person_, then the tidy dataset has one pair of values per person, and one row for each person. 
We can use the handy `tidyr::uncount()` function, the opposite of `dplyr::count()`:

```{r}
haireye %>% 
  tidyr::uncount(n) %>% 
  DT::datatable(rownames = FALSE)
```


### Untidy Examples

The following are examples of untidy data. 
They're untidy for either of the cases considered above, but for discussion, let's take a _hair-eye color combination_ to be one observational unit. 

Note that untidy does not always mean "bad", just inconvenient for the analysis you want to do.

__Untidy Example 1__: 
The following table is untidy because there are multiple observations per row. 
It's _too wide_.

Imagine calculating the total number of people with each hair color. 
You can't just `group_by()` and `summarize()`, here!

```{r, echo = FALSE}
haireye_untidy <- haireye %>% 
  mutate(eye = str_c(eye, "_eyed")) %>% 
  pivot_wider(id_cols = hair, names_from = eye, values_from = n)
knitr::kable(haireye_untidy)
```

This sort of table is common when presenting results.
It's easy for humans to read, but hard for computers to work with.
Untidy data is usually that way because it was structured for human, not machine, reading.


__Untidy Example 2__: 
The following table is untidy for the same reason as Example 1—multiple observations are contained per row. 
It's _too wide_.

```{r, echo = FALSE}
haireye %>% 
  mutate(hair = str_c(hair, "_haired")) %>% 
  pivot_wider(id_cols = eye, names_from = hair, values_from = n) %>% 
  knitr::kable()
```


__Untidy Example 3__: 
This is untidy because each observational unit is spread across multiple columns. 
It's _too long_. 

In fact, we needed to add an identifier for each observation, otherwise we would have lost which row belongs to which observation! 

Does red hair ever occur with blue eyes? 
Can't just `filter(hair == "red", eye == "blue")`!

```{r, echo = FALSE}
haireye %>% 
  mutate(.id = 1:n()) %>% 
  pivot_longer(cols = hair:eye, names_to = "body_part", values_to = "color") %>% 
  select(-n, n) %>% 
  DT::datatable(rownames = FALSE)
```


__Untidy Example 4__: 
Just when you thought a data set couldn't get any longer! 
Now, each variable has its own row: hair color, eye color, and `n`. 


```{r, echo = FALSE}
haireye %>% 
  mutate(obs = 1:n(),
         n   = as.character(n)) %>% 
  pivot_longer(cols = c(hair, eye, n), names_to = "variable", values_to = "value") %>% 
  DT::datatable(rownames = FALSE)
```

_This is the sort of format that is common pulling data from the web or other "Big Data" sources._


### Pivoting tools

The task of making tidy data is about making data either _longer_, by stacking two or more rows, or _wider_, by putting one or more columns alongside each other based on groups. 
This is called __pivoting__.

Usually the task of tidying data involves lengthening, and usually the task of widening is useful for turning data into something more friendly for human eyes.

Sometimes, you will see data described as being in "long" or "wide" formats.
Those terms aren't that useful—"long" and "wide" are relative terms. 

The easiest and most powerful way to widen or lengthen data are with the functions `tidyr::pivot_wider()` and `tidyr::pivot_longer()`.

History: R has seen many attempts at reshaping that have progressively gotten better. 

First came the `reshape` and `reshape2` packages. 

  - Both were finicky. 
  - Used function names that I could never remember: `melt()` and `cast()`. 
  
Then, `tidyr` package replaced these.

  - The `tidyr::spread()` and `tidyr::gather()` functions provided a simple interface.
  - I still couldn't remember these names. 
  - Still a little hard to use.
  
We will use `tidyr::pivot_longer()` and `tidyr::pivot_wider()`.


## Univariate Pivoting

Let's start with pivoting in the simplest case where only one variable is "out of place". 
We'll use the hair and eye color example from before, using the untidy data version from Example 1:

```{r}
haireye_untidy <- haireye %>% 
  mutate(eye = str_c(eye, "_eyed")) %>% 
  pivot_wider(id_cols = hair, names_from = eye, values_from = n)

haireye_untidy
```

The _eye color_ variable is spread out across columns. 
To fix this, we need to convert the eye color columns to two columns:

- one column to hold the eye color (column names),
- one column to hold the values.

Doing this, we obtain:

```{r, echo = FALSE}
haireye_untidy %>% 
  pivot_longer(contains("eyed"), names_to = "eye", values_to = "n")
```

Let's dig into that!

### `pivot_longer()`

`pivot_longer()` takes a data frame, and returns a data frame. 

The main arguments after the data argument that we'll need are:

- `cols`: the columns we want to pivot into a single column. Give the column names
- `names_to`: the old column _names_ are going to be stored in a new column. 
  What should this new column be named?
- `values_to`: the _values_ in the old columns are going to stored in a new column.
  What should this new column be named?

Possibly the trickiest bit is in identifying the column names. 
We could list all of them:

```{r}
haireye_untidy %>% 
  pivot_longer(cols      = c(Blue_eyed, Brown_eyed, Green_eyed, Hazel_eyed),
               names_to  = "eye",
               values_to = "n")
```

That can be a little tedious.

We could identify a range. This is efficient, but not so robust if the data changes.

```{r}
haireye_untidy %>% 
  pivot_longer(cols      = Blue_eyed:Hazel_eyed,
               names_to  = "eye",
               values_to = "n")
```

Better is to use helper functions from the `tidyselect` package. 
In this case, we know the columns contain the text "eyed", so let's use `tidyselect::contains()`:

```{r}
haireye_untidy %>% 
  pivot_longer(cols      = contains("eyed"),
               names_to  = "eye",
               values_to = "n")
```

Yet another way is to indicate everything except the `hair` column:

```{r}
haireye_untidy %>% 
  pivot_longer(cols      = -hair,
               names_to  = "eye",
               values_to = "n")
```


### `pivot_wider()`

Let's say we want go from a longer data frame to a shorter data frame.
This is the opposite of what we did above. 

We might want to do this:

  - To make a table for presentation
  - With longitudinal or family data, to to go from **multilevel models** (which need longer data; each row is an *observation*) to **SEM analyses** (which need wider data; each row is an *individual* or *family*)
  
For example, if we want to go from:

```{r, echo = FALSE}
haireye_untidy %>% 
  pivot_longer(contains("eyed"), names_to = "eye", values_to = "n")
```

To: 

```{r, echo = FALSE}
haireye_untidy
```

We need to: 

  - Take the column `eye` and make *each unique entry* a new column
  - Take the column `n` and make these values in the new `eye` columns.

`pivot_wider()` is the reverse of `pivot_longer()`.

Like `pivot_longer()`, `pivot_wider()` takes a data frame and returns a data frame. 
The main arguments after the data argument that we'll need are:

- `id_cols`: The columns you would like to **keep**. 
  For example, the identifier number for the observation.
- `names_from`: The new column names are coming from an old column. 
  Which column is this?
- `values_from`: The column values are coming from an old column. 
  Which column is this?

```{r}
haireye %>% 
  pivot_wider(id_cols     = hair, 
              names_from  = eye, 
              values_from = n)
```


### Activity

Fill out the [__Univariate Pivoting Exercise__](https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/tutorials/s06_pivot-exercise.Rmd).


## Multivariate Pivoting

Now let's consider the case when more than one variable are "out of place".
Perhaps there are multiple variables per row and/or multiple observations per row.

For example, consider the (lightly modified) `iris` data set that we'll call `iris2`:

```{r, echo = FALSE}
iris2 <- iris %>%
  mutate(id = 1:n()) %>% 
  rename(species = Species) %>% 
  pivot_longer(c(-species, -id), 
               names_to  = "variable", 
               values_to = "measurement") %>% 
  mutate(variable = variable %>% 
           str_replace("\\.", "_") %>% 
           tolower()) %>% 
  pivot_wider(c(id, species), 
              names_from  = variable, 
              values_from = measurement)
DT::datatable(iris2, rownames = FALSE)
```

Although we probably wouldn't, we could view this as having _two variables bundled into the column names_:

- "Plant part", either `sepal` or `petal`.
- "Dimension", either `length` or `width`.

The resulting tidy data frame would then be:

```{r, echo = FALSE}
iris2_longest <- iris2 %>% 
  pivot_longer(cols      = c(-species, -id), 
               names_to  = c("part", "dimension"),
               names_sep = "_",
               values_to = "measurement")
```

```{r}
iris2_longest
```

More realistic is the situation where there are _multiple observations per row_:

- An observation of (length, width) of the sepal.
- An observation of (length, width) of the petal.

The resulting tidy data frame has a length that's in between the above two:

```{r, echo = FALSE}
iris2_longer <- iris2 %>% 
  pivot_longer(cols      = c(-id, -species), 
               names_to  = c("part", ".value"), 
               names_sep = "_")
```

```{r}
iris2_longer
```


### `pivot_longer()`

To obtain the case where two (or more) variables are contained in column names, here's how we specify the arguments of `pivot_longer()`:

- `cols`: As usual.
- `names_sep`: What is separating the variables in the column names?
- `names_to`: The old columns are going to be put into new columns, after being separated. What should those columns be named?
- `values_to`: As usual. 

Here is the code:

```{r}
iris2 %>% 
  pivot_longer(cols      = c(-species, -id), 
               names_to  = c("part", "dimension"),
               names_sep = "_",
               values_to = "measurement")
```

To obtain the case where multiple observations are contained in one row, here's how to specify the arguments of `pivot_longer()`:

- `cols`: As usual.
- `names_sep`: As above.
- `names_to`: As above, except this time, one part of the old column names are going to stay as columns (in this case, "length" and "width"). Indicate `".value"` instead of a new column name.  
- `values_to`: Not needed! You've already indicated that using the `".value"` placeholder. 
 
```{r}
iris2 %>% 
  pivot_longer(cols      = c(-id, -species), 
               names_to  = c("part", ".value"), 
               names_sep = "_")
```


### `pivot_wider()`

If two or more columns contain parts of a variable name (i.e., each unique combination of these columns gives rise to a new variable), here's how we can use `pivot_wider()`:

- `id_cols`: as usual.
- `names_from`: the new variable names are coming from old columns. Which old columns?
- `names_sep`: What character should you separate the entries of the old columns by?
- `values_from`: as usual.

Here is the code to go from the longest form to the original:

```{r}
iris2_longest %>% 
  pivot_wider(id_cols     = c(id, species),
              names_from  = c(part, dimension), 
              names_sep   = "_", 
              values_from = measurement)
```

If variables are spread out amongst rows _and_ columns (for example, "sepal width" has "sepal" in a column, and "width" as a column name), here's how we can use `pivot_wider()`:

- `id_cols`: as usual
- `names_from`: Which column contains the part of the variable?
- `names_sep`: As before, what character should you separate the entries of the old columns by?
- `values_from`: Which column names contain the other part of the variable?

Here is the code to go from the "semi-long" form to the original:

```{r}
iris2_longer %>% 
  pivot_wider(id_cols     = c(id, species), 
              names_from  = part, 
              names_sep   = "_",
              values_from = c(length, width))
```


### Activity

Fill out __Exercise 2: Multivariate Pivoting__ in the 
[worksheet](https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/tutorials/s06_pivot-exercise.Rmd).


## Making tibbles

In base R, we can make data frames using the `data.frame()` function. The tidyverse version is `tibble::tibble()`, which also has backwards referencing to variables you make on the fly. It's also stricter by not allowing recycling unless the vector is of length 1:

Good:

```{r}
tibble(x = 1:6,
       y = min(x))
```

Bad:

```{r, error = TRUE}
tibble(x = 1:6,
       y = 1:2)
```

Truly manual construction of tibbles is easy with `tibble::tribble()`:

```{r}
tribble(
  ~Day, ~Breakfast,
  1, "Apple",
  2, "Yogurt",
  3, "Yogurt"
)
```

Check out the `datapasta` package for ways to reproducibly copy-paste data from 
spreadsheets into R. `datapasta` uses `tribble()`.

List columns are easy with tibbles!

```{r}
(list_col <- tibble(n = 1:2,
                    y = list(iris, mtcars)))
```

Often obtained with `nest()` and `unnest()`:

```{r}
(iris_nest <- iris %>% 
  group_by(Species) %>% 
  nest())
```

```{r}
iris_nest %>% 
  unnest(data)
```

`expand_grid()` to obtain all combinations:

```{r}
expand_grid(x = 1:2, y = 1:2, z = 1:2)
```

In conjunction with `nesting()`:

```{r}
expand_grid(nesting(x = 1:2, y = 1:2), z = 1:2)
```


## Implicit `NA`'s

Sometimes there's "hidden" missing data in a tibble. Here's an example  
from the documentation of `tidyr::expand()`:

```{r}
(df <- tibble(
  year   = c(2010, 2010, 2010, 2010, 2012, 2012, 2012),
  qtr    = c(   1,    2,    3,    4,    1,    2,    3),
  return = rnorm(7)
))
```

Here, there are values for `qtr` 1, 2, 3, 4 in `year` 2010 and 
`qtr` 1, 2, 3 in `year` 2012, but no `qtr` 4 in `year` 2012. 
The value of `return` in `year` 2012, `qtr` 4 is "implicilty" `NA`.

We can use functions to fill out this data frame and make these "implicit" missing
values explicit.

We can consider all existing combinations by invoking the column names in 
`expand()` or `complete()` (which either _drops_ or _keeps_ all other columns):

```{r}
df %>% 
  expand(year, qtr)
df %>% 
  complete(year, qtr)
```

We can consider new combinations by specifying an expectation of possible values:

```{r}
df %>% 
  expand(year = 2010:2012, qtr)
df %>% 
  complete(year = 2010:2012, qtr)
```

Want to link two or more columns when looking for combinations? Use `nesting()`.


## Activity (10 min)

Fill out __Exercise 3: Making tibbles__ in the 
[worksheet](https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/tutorials/s06_pivot-exercise.Rmd).
