---
editor_options: 
  markdown: 
    wrap: 72
---

# File input/output (I/O)

Today's class is all about reading into R and writing out data files out
of R.

Because sometimes you want to work with your own damn data, instead of
data about penguins, cars, life expectancy, or the Big Five personality
scores of 2800 of your closest friends.

## Resources

### References and tutorials

-   Jenny Bryan's [notes](https://stat545.com/import-export.html) on
    file I/O
-   [Tutorial](https://beanumber.github.io/sds192/lab-import.html#data_from_an_excel_file)
    on importing Excel files
-   [Tutorial](http://jenrichmond.rbind.io/post/how-to-use-the-here-package/)
    on relative paths and how RStudio treats .R script files and .Rmd
    files

### Package documentation

-   [readr::read_csv()](https://readr.tidyverse.org/reference/read_delim.html)
-   [readxl::read_excel()](https://readxl.tidyverse.org/reference/read_excel.html)
-   [googlesheets4::read_sheet()](https://googlesheets4.tidyverse.org/reference/read_sheet.html)
-   [haven](https://haven.tidyverse.org/) package for importing SPSS,
    SAS, Stata files
-   [rio::import()](https://cloud.r-project.org/web/packages/rio/vignettes/rio.html) for importing all manner of formats with one command!
-   [here::here()](https://here.r-lib.org)

<!-- The following chunk allows errors when knitting -->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

## Writing data to disk

Let's first load the built-in gapminder dataset and the tidyverse:

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(gapminder)
library(tidyverse)
```

Next, let's filter the data to only 2007 and only the Asia continent and
save it to a new object.

```{r}
(gap_asia_2007 <- gapminder %>% filter(year == 2007, continent == "Asia"))
```

We can write this to a comma-separated value (csv) file with the command:

```{r eval=FALSE}
write_csv(gap_asia_2007, "exported_file.csv")
```

But where did this file go? 
We should save the file in a sensible location. 
We need to practice controlling where R is running, where it looks for files, and where it writes out your files. 
To that end, let's review the working directory and RStudio Projects.


## The working directory and RStudio Projects

When you open R, it "runs" in some folder on your computer. 
This is the place it will look for files to import and write files as output.
Think about where your participation and homework files end up when you knit them.

If you have R/RStudio closed, and you open a .R or .Rmd file, R/RStudio will start in the folder holding that file.

If you open R/RStudio from the Windows Start menu, the Mac dock, the Mac Spotlight, etc., R/Studio will start in its default location (probably your user home directory, see Tools → Global Options → General → Default working directory...).

Write all of your R scripts assuming:

1.  You are running them by opening a fresh new R session

    -   Don't use `rm(list = ls())` to clean the workspace--the workspace is already clean
    -   You need to load required pacakges with `library()`
    -   Don't work on several different projects in one R session at the same time!

2.  You have all of the necessary packages installed

    -   Don't include `install.packages()` calls unless they are commented out or otherwise set not to run

3.  The script will run without human input

    -   You need to import or load any data you are working with
    -   Load data and write output using R commands, not `file.choose()`, `read.clipboard()`, the buttons in RStudio, etc.

4.  All of the needed files live in your project folder

    -   Write *relative* paths, rather than *absolute* paths
    -   `file.path("data", "cats-data_2020-03-04.xlsx")`
    -   Not:
        `/Users/brenton/Research/cats_project/data/cats-data_2020-03-04.xlsx`
        or
        `C:\\Users\\brenton\\Research\\cats_project\\data\\cats-data_2020-03-04.xlsx`\
        or`file.path("C:", "Users", "brenton", "Research", "cats_project", "data", "cats-data_2020-03-04.xlsx")`

5.  Your script might run on any system

    -   Write the paths to files using `file.path()` or `here::here()`, rather than typing out the whole path 
    -   This is because Windows and Mac/Linux have different syntaxes for file paths:

        -   Mac/Linux: `path/to/folder`
        -   Windows: `path\\to\\folder`

This approach has several advantages:

1.  Frictionless running on different computers
2.  Less breakage (e.g., if you move a folder around, the relative paths will still work)
3.  Less surprising or weird behaviors due to session crud
4.  Easy to tweak data/code and update results

### RStudio projects

RStudio projects can help with following these best practices. 

The best way to work with R is to have a separate Project for each project/paper you are working on. Switch RStudio to that Project whenever you are working on those files. 

You can you are working in a Project because its name will be in the RStudio title bar. 
One macOS, the Project name will also show on the Dock icon.


## here::here()

The *here* package is the best way to refer to files. 

The major function is `here::here()`. 
Like `file.path()`, `here::here()` lets you specify a path to a file and then adds the system-appropriate separators (`/` or `\\`). 

Where `here::here()` shines is that it figures out where the relative paths should start from. 
It looks around in the folders in your directory and finds the .Rproj file, then constructs the relative file paths from there.

The .Rproj file is a sign post that says "START HERE".

For example, create a new folder "data" in your participation folder. 
Then, save `gap_asia_2007` using the `here::here()` and `write_csv()` functions:

```{r, eval = FALSE}
write_csv(gap_asia_2007, 
          here::here("participation", "data", "gap_asia_2007.csv")
)
```

More details on `here` are available in this [short
article](http://jenrichmond.rbind.io/post/how-to-use-the-here-package/).


## Reading data from disk

The same csv file that we just saved to disk can be imported into R again by specifying the path where it exists:

```{r, eval = FALSE}
dat <- read_csv(here::here("participation", "data", "gap_asia_2007.csv"))
dat
```

Notice that the output of the imported file is the same as the original
tibble.
`read_csv()` was intelligent enough to detect the types of the columns. 
This won't always be true so it's worth checking! 
In particular, be on the lookout for any columns it imports as `col_character()`!

The [read_csv()](https://readr.tidyverse.org/reference/read_delim.html) function has many additional options including the ability to specify column types (e.g., is "1990" a year or a number?), skip columns, skip rows, rename columns on import, trim whitespace, and more.

To control the column types, use the `cols()` function:

```{r, eval = FALSE}
dat <- read_csv(
  here::here("participation", "data", "gap_asia_2007.csv"),
  col_types = cols(
    country = col_factor(),
    continent = col_factor(),
    year = col_date(format = "%Y"),
    .default = col_double() # all other columns as numeric (double)
  )
)
dat
```

By default, it leaves all columns as `col_guess()`, but it's better to be explicit.

Another important option to set is the `na` argument, which specifies what values to treat as `NA` on import. 
By default, `read_csv()` treats blank cells (i.e., `""`) and cells with `"NA"` as missing. 
You might need to change this (e.g., if missing values are entered as `-999`). Note that `readxl::read_excel()` by default only has `na = c("")` (no `"NA"`)!

```{r, eval = FALSE}
dat <- read_csv(
  here::here("participation", "data", "gap_asia_2007.csv"),
  col_types = cols(
    country = col_factor(),
    continent = col_factor(),
    year = col_date(format = "%Y"),
    .default = col_double() # all other columns as numeric (double)
  ),
  na = c("", "NA", -99, "No response")
)
dat
```

## Import a file from the web/cloud

### Import a CSV file from the internet

To import a CSV file from a web, assign the URL to a variable

```{r}
url <- "http://gattonweb.uky.edu/sheather/book/docs/datasets/magazines.csv"
```

and then apply read_csv file to the `url`.

```{r}
dat <- read_csv(url)
```

You can do this in one step if you like:

```{r eval = FALSE}
read_csv("http://gattonweb.uky.edu/sheather/book/docs/datasets/magazines.csv")
```

(This doesn't work for all import functions, so I usually keep them separate as two steps.)

### Import an Excel file (.xls or .xlsx) from the internet

First, we'll need the package to load in Excel files:

```{r}
library(readxl) 
```

Datafiles from this tutorial were obtained from:
<https://beanumber.github.io/sds192/lab-import.html#data_from_an_excel_file>

Unlike with a CSV file, to import an .xls or .xlsx file from the internet, you first need to download it locally.

*Note: The folder you want to save the file to _has to exist_!.*
*If it doesn't, _you will get an error._*

You can create the folder path in one of three ways:

  1. Create them directly in Finder/Windows Explorer
  2. Use the buttons in the Files tab in RStudio
  3. Use the `dir.create()` function:

```{r, eval = FALSE}
if ( !dir.exists( here::here("participation", "data") ) ) {
  dir.create( here::here("participation", "data"), recursive = TRUE )
}
```

Next, you download the file.
To download it, create a new object called `xls_url` and then use `download.file()` to download it to a specified path.

```{r, eval = FALSE}
xls_url <- "http://gattonweb.uky.edu/sheather/book/docs/datasets/GreatestGivers.xls"
download.file(
  xls_url, 
  here::here("participation", "data", "some_file.xls"), 
  mode = "wb"
)
```

**NOTE:** Don't assign the result of `download.file()`.

**NOTE:** The `mode = "wb"` argument at the end is really important if you are on Windows. 
If you omit, you will probably get a message about downloading a corrupt file.
More details about this behavior can be found [here](https://github.com/HenrikBengtsson/Wishlist-for-R/issues/30).


Naming a file "some_file" is extremely bad practice (hard to keep track of the files).
You should **always** give it a more descriptive name. 
It's also a good idea to avoid spaces in filenames.
You should come up with a system for naming your files and use it consistently. 
My file names look like this: `progdata_example-dataset_2021-03-02_bmw.csv`


Often, it's a good idea to name the file similarly (or the same as) the original file (sometimes that might not be a good idea if the original name is non-descriptive).

There's handy trick to extract the filename from the URL:

```{r, eval = FALSE}
file_name <- basename(xls_url)
download.file(
  xls_url, 
  here::here("participation", "data", file_name), 
  mode = "wb"
)
```

Now we can import the file:

```{r, eval = FALSE}
dat <- read_excel(
  here::here("participation", "data", file_name)
)
```


## Read in a sample SPSS file.

Let's load a sample SPSS file and work with it. 
Download the file from [here](https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/data/clevel.sav) and save it in your `participation/data` folder.

These data are a random subset of the data used in [this paper](https://doi.org/10.1016/j.jvb.2018.02.005). 
This was a study looking at personality traits that distinguish C-level executives from lower-level managers among men and women.

The subset of data here consists of 200 cases, with variables indicating: 

  1. The language of assessment (English, Dutch, or French) 
  2. Gender 
  3. C-level or not 
  4. Extraversion level, as well as 4 facet traits (Leading, Communion, Persuasive, Motivating)

Let's load in the data using the `haven` package.

```{r, eval = FALSE}
(clevel <- haven::read_spss(
  here::here("participation", "data", "clevel.sav")
)
)
```

Notice that this tibble looks a little different for the `language` and `gender` variables than normal. 
It has **labels** for the numeric values. 
This format is what SPSS uses, but it's not standard for R.
Let's convert those variables, and `isClevel` as well, to factors:

```{r, eval = FALSE}
clevel_cleaned <-
  clevel %>% 
  mutate(language = as_factor(language),
         gender = as_factor(gender),
         isClevel = factor(isClevel, 
                           levels = c(0, 1), 
                           labels = c("No", "Yes"))
  )
print(clevel_cleaned)
```

Notice how the variables are now factors with labels as the entries, instead of the original code numbers.

## Saving Files

### Saving data frames

Let's save this file as a CSV file so that it's a smaller file and easier to import again in the future.

```{r, eval = FALSE}
write_csv(
  clevel_cleaned, 
  here::here("participation", "data", "clevel_cleaned.csv")
)
```

### Saving plots

Now let's make a plot.

```{r, eval = FALSE}
clevel_plot <-
  clevel_cleaned %>% 
  mutate(isClevel = recode(isClevel, 
                           No = "Below C-level", 
                           Yes = "C-level"),
         gender = recode(gender,
                         Female = "Women",
                         Male = "Men")) %>% 
  ggplot(aes(paste(isClevel, gender, sep = "\n"), Extraversion, color = gender)) +
  geom_boxplot() +
  geom_jitter(height = .2) +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  ggtitle("Extraversion Stan Scores") +
  scale_y_continuous(breaks = 1:9) +
  ggthemes::theme_fivethirtyeight()

clevel_plot
```

Let's save the plot in several formats. 
This is useful if we want to use the plot outside of Markdown. 
Save plots using the `ggsave()` funnction.

`ggsave()` has many options. 
See the help function `?ggsave` for full details. 
The main arguments are:

  - `filename`
  - `plot`
  - `width` and `height` (inches by default)
  - `dpi` (dots per inch; for bitmap formats).

`ggsave()` will try to guess what format you want based on the file name. 
If you want, you can specify a specific format or R graphics device to save with using the `device` argument.

```{r, eval = FALSE}
dir.create(here::here("participation", "output", "figures"), recursive = TRUE)
ggsave(here::here("participation", "output", "figures", "clevel_extraversion.svg"), 
       clevel_plot,
       height = 6,
       width = 6
)
```

You can save to several formats. 
Generally, work with a *vector* format like `.svg`, `.eps`, or `.pdf`. 

Vector graphics represent the image as a series of data points and equations. 
This means that they can be made smaller or larger or zoomed in on without damaging the image quality.

If you can't use a vector format for some reason, you can also export to a *bitmap* format. 
Bitmap graphs represent the image as colored dots or pixels. 
This means that the image quality will suffer if you make the image larger or zoom in on it (making it smaller can also sometimes compromise quality). 
With bitmap images, you need to be concerned with *resolution* (how many pixels/dots per inch when printed). 
Always use at least **300 DPI** resolution.

There are several popular bitmap image formats. 

  - `.tiff`/`.tif` 
    - Highest quality, but also the largest. 
    - Use it for print graphics, but you should probably avoid it for images to be hosted on the web. 
  - `.png` 
    - A bit smaller, and it should be your go to for charts, figures, line drawings, etc. 
    - More complex images (e.g., photos) can get pretty big with `.png`, though. 
  - `.jpg`/`.jpeg`
    - Probably the most popular bitmap format.
    - Works well for photographs hosted on the web, but its compression often makes line drawing and charts look terrible.
    - `.jpg`/`.jpeg` also degrades in quality each time you edit/save it, so don't use it for images you intend to edit. 
  - `.gif`
    - Generally avoid `.gif` unless you are making an animation or you need very small file size for a simple image. 
    - `.gif` supports very few colors, so always check your image quality after making a `.gif`.

Let's save to some other formats:

```{r, eval = FALSE}
ggsave(here::here("participation", "output", "figures", "clevel_extraversion.pdf"), clevel_plot, height = 6, width = 4)
ggsave(here::here("participation", "output", "figures", "clevel_extraversion.tiff"), clevel_plot)
ggsave(here::here("participation", "output", "figures", "clevel_extraversion.png"), clevel_plot)
```


## Organizing Your Project Folders

Follow a consistent folder structure for all of your projects. 
This will make it easier for you to say organized and make your code, data, and projects easy to share.

In your project's root folder, you should have a README.md file and a .Rproj project. 
Then, you should have folders that separate different types of files:

-   `data`: Stores all of your data files for a project

    -   Have subfolders for different dates, waves, groups, etc. as needed
    -   If you plan on saving cleaned data, having separate `data_raw` and `data` folders is a good idea.

-   `output`: Stores any output your scripts generate

    -   Depending on how many figures and other output files you will create, you might want to split/subfolder this into `figures`, `reports`, etc.

You might add additional folders, such as:

-   `tests`: A folder that includes tests to check that your scripts or results are accurate. Check out the `testthat` package.

-   `templates`: A folder to hold template files (e.g., RMarkdown templates, Word templates, CSS files for HTML output, TeX templates for PDF output)

-   `admin`: For administrative documents (e.g., IRB approval, grant information)

-   `doc`: For documentation (e.g., variable codebooks, style guides)

-   `R` or `src` or `scripts`: Folders to store functions and scripts that you call from your markdown (e.g., a data import and cleaning script)

    -   Use folders or subfolders for each language if you are programming in multiple languages: `R`, `python`, `sql`, `C`, etc.
    -   Can put them all in a generic `scripts` or `src` folder if not too many
    
-   `markdown`: If you have multiple RMarkdown documents for your project, consider putting them into a subfolder


## Bonus Activity: `rio`

Do the `clevel` activity again, but this time use the `rio::import()` and `rio::export()` functions instead of the `read_*()` and `write_*()` functions. 
In `rio::import()`, be sure to specify: `rio::import(..., setclass = "tibble")`.


## Entering Data Manually

### Making data.frames or tibbles

In base R, we can make data frames using the `data.frame()` function. 
The tidyverse version is `tibble::tibble()`. `tibble()` is stricter by not allowing recycling unless the vector is of length 1.

Good:

```{r}
dat <- tibble(
  x = 1:6,
  y = min(x)
)
```

Bad:

```{r, error = TRUE}
dat <- tibble(
  x = 1:6,
  y = 1:2
)
```


Manual construction of tibbles is convenient with `tibble::tribble()`:

```{r}
dat <- tribble(
  ~Day, ~Breakfast,
  1, "Apple",
  2, "Yogurt",
  3, "Yogurt"
)
```

### datapasta

The *datapasta* package helps to reproducibly copy-paste data from spreadsheets into R. 
`datapasta` uses `tribble()`.

```{r, eval = FALSE}
install.packages("datapasta")
```

After you install the package, you can set keyboard shortcuts in RStudio for its functions:

  1. Click Tools → Modify Keyboard Shortcuts…
  2. Search for 'paste'.
  3. Set keyboard shortcuts for "Paste as tribble", "Paste as vector", and other functions if you like.

Now, try out pasting a vector and a tibble.
