# File input/output (I/O)

Today's class is all about reading into R and writing out data files out of R.

Because sometimes you want to work with your own damn data, instead of data 
about flowers, cars, life expectancy, or the Big Five personality scores of 2800
of your closest friends.

## Worksheet

Normally there would be a separate worksheet file, but we'll do everything 
in-line in the guide for this class.

## Resources

### References and tutorials

* Jenny Bryan's [notes](https://stat545.com/import-export.html) on file I/O
* [Tutorial](https://beanumber.github.io/sds192/lab-import.html#data_from_an_excel_file) 
  on importing Excel files
* [Tutorial](http://jenrichmond.rbind.io/post/how-to-use-the-here-package/) on 
  relative paths and how RStudio treats .R script files and .Rmd files

### Package documentation

* [readr::read_csv()](https://readr.tidyverse.org/reference/read_delim.html)
* [readxl::read_excel()](https://readxl.tidyverse.org/reference/read_excel.html)
* [googlesheets4::read_sheet()](https://googlesheets4.tidyverse.org/reference/read_sheet.html)
* [haven](https://haven.tidyverse.org/) package for importing SPSS, SAS, Stata files
* [rio::import()](https://cloud.r-project.org/web/packages/rio/vignettes/rio.html) for importing all manner of formats with one command!
* [here::here()](https://here.r-lib.org)

<!-- The following chunk allows errors when knitting -->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

## Writing data to disk

Let's first load the built-in gapminder dataset and the tidyverse:

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(gapminder)
library(tidyverse)
```

Next, let's filter the data to only 2007 and only the Asia continent and 
save it to a new object.

```{r}
(gap_asia_2007 <- gapminder %>% filter(year == 2007, continent == "Asia"))
```

We can write this to a comma-separated value (csv) file with just one command:

```
write_csv(gap_asia_2007, "exported_file.csv")
```

But where did this file go? We should save the file in a sensible location. 
We need to practice controlling where R is running, where it looks for files, 
and where it writes out your files. To that end, let's talk about the working 
directory and RStudio Projects.

## The working directory and RStudio Projects

When you open, it "runs" in some folder on your computer. This is the place
it will look for files to import and write files as output. Think about where
your participation and homework files end up when you knit them.

If you have R/RStudio closed, and you open a .R or .Rmd file, R/RStudio will start
in the folder holding that file.

If you open R/RStudio from the Windows Start menu, the Mac dock, the Mac Spotlight,
etc., R/Studio will start in its default location (probably your user home directory,
see Tools → Global Options → General → Default working directory…).

### The working directory

When I say "R/Studio will start in…", what I am referring to is R's "working 
directory". Like I say above, this is the place R will look for files to import 
and write files as output. You can check what R's current working directory 
is using the `getwd()` function:

```{r}
getwd()
```

You can also change the working directory using the `setwd()` function:

```{r, eval = FALSE}
setwd(file.path("path", "to", "folder"))
```

Do not use `setwd()`! You should always write your R scripts so that the entire
project is self-contained in a folder. All of the scripts, folders, data, output, 
etc. should all "live" within this project folder. 

Write all of your R scripts assuming:

  1. You are running them by opening a fresh new R session
     - Don't use `rm(list = ls())` to clean the workspace--the workspace is already clean
     - You need to load required pacakges with `library()`
     - Don't work on several different projects in one R session at the same time!
  2. You have all of the necessary packages installed
     - Don't include `install.packages()` calls unless they are commented out
       or otherwise set not to run
  3. The script will run without human input
     - You need to import or load any data you are working with
     - Load data and write output using R commands, not `file.choose()`, 
       `read.clipboard()`, the buttons in RStudio, etc.
  4. All of the needed files live in your project folder
     - Write _relative_ paths, rather than _absolute_ paths
     - `data/cats-data_2020-03-04.xlsx` or `file.path("data", "cats-data_2020-03-04.xlsx")`
     - Not: `/Users/brenton/Research/cats_project/data/cats-data_2020-03-04.xlsx` 
       or `C:\\Users\\brenton\\Research\\cats_project\\data\\cats-data_2020-03-04.xlsx`  
       or`file.path("C:", "Users", "brenton", "Research", "cats_project", "data", "cats-data_2020-03-04.xlsx")`
  5. Your script might run on any system
     - Write the paths to files using `file.path()` or `here::here()`, rather
       than typing out the whole path (Windows and Mac/Linux have different 
       syntaxes for file paths):
       - Mac/Linux: `path/to/folder`
       - Windows: `path\\to\\folder`

This approach has several advantages:

  1. Frictionless running on different computers
  2. Less breakage (e.g., if you move a folder around, the relative paths will still work)
  3. Less surprising or weird behaviors due to session crud
  4. Easy to tweak data/code and update results


### RStudio projects

RStudio projects can help with following these best practices. Let's make a project.

Click `File → New Project…` and then make a new project in your participation repo
folder. Give it a useful name like `DataSci-participation.Rproj`.

When you double click on a .Rproj file, it:
  1. Opens a new fresh R session, with
  2. The working directory set to the location of the .Rproj file, and
  3. No connection whatsoever to any other R sessions you already have open
  
You can also set specfic options for each RStudio project (e.g., number of spaces
to insert when you type Tab, etc.).

Let's practice closing RStudio and re-opening it by opening the RStudio project.
Run `getwd()` to see where R is running. 


## here::here()

You can make your projects very portable by using .Rproj RStudio projects and 
relative paths with `file.path()`. This still can be a little fiddly:

  1. Need to remember to open the .Rproj file, not a .Rmd or .R file
  2. If you open a .Rmd file contained in a subfolder, the working directory
     won't be right for working with `file.choose()` for relative paths.
     
Enter the `here` package. `here` takes the idea of relative paths and extends it
in very useful ways. 

The major function is `here::here()`. Like `file.path()`, `here::here()` lets you
specify a path to a file and then adds the system-appropriate separators (`/` or `\\`).
Where `here::here()` shines is that it figures out better where the relative paths
should start from. It looks round in the folders in your directory and finds
the .Rproj file, then constructs the relative file paths from there.

For example, create a new folder "data" in your participation repo, and within 
it make a subfolder called "s008_data". Then, save `gap_asia_2007` using the
`here::here()` and `write_csv()` functions:

```{r, eval = FALSE}
write_csv(gap_asia_2007, here::here("data", "s008_data", "exported_file.csv"))
```

More details on `here` are available in this 
[short article](http://jenrichmond.rbind.io/post/how-to-use-the-here-package/).


## Reading data from disk

The same csv file that we just saved to disk can be imported into R again by 
specifying the path where it exists:

```{r, eval = FALSE}
read_csv(here::here("data", "s008_data", "exported_file.csv"))
```

Normally we would store the imported data into a new object that you can use in
subsequent analyses using the assignment function `<-`. 

Notice that the output of the imported file is the same as the original tibble, 
and `read_csv()` was intelligent enough to detect the types of the columns. 
This won't always be true so it's worth checking! The 
[read_csv package](https://readr.tidyverse.org/reference/read_delim.html) has 
many additional options including the ability to specify column types (e.g., 
is "1990" a year or a number?), skip columns, skip rows, rename columns on 
import, trim whitespace, and more.

One of the most important options to set is the `na` argument, which specifies 
what values to treat as `NA` on import. By default, `read_csv()` treats blank 
cells (i.e., `""`) and cells with `"NA"` as missing. You might need to change
this (e.g., if missing values are entered as `-999`). Note that `readxl::read_excel()` 
by default only has `na = c("")` (no `"NA"`)!


## Import a file from the web/cloud

### Import a CSV file from the internet

To import a CSV file from a web, assign the URL to a variable

```{r}
url <- "http://gattonweb.uky.edu/sheather/book/docs/datasets/magazines.csv"
```

and then apply read_csv file to the `url`.

```{r}
read_csv(url)
```

You can do this in one step if you like:

```
read_csv("http://gattonweb.uky.edu/sheather/book/docs/datasets/magazines.csv")
```

(This doesn't work for all import functions, so I usually keep them separate
as two steps.)


### Import an Excel file (.xls or .xlsx) from the internet

First, we'll need the package to load in Excel files:

```{r}
library(readxl) 
```

Datafiles from this tutorial were obtained from:
https://beanumber.github.io/sds192/lab-import.html#data_from_an_excel_file

Unlike with a CSV file, tto import an .xls or .xlsx file from the internet, you 
first need to download it localy. 

**Note: The folder you want to save the file to _has to exist_!. If it doesn't,**
**you will get an error.**
  
Either create the folder path in Finder/Windows Explorer, or use the `dir.create()`
function:

```{r, eval = FALSE}
dir.create(here::here("data", "s008_data"), recursive = TRUE)
```

Next, you download the file. To download it, create a new object called `xls_url` 
and then using `download.file` to dowload it to a specified destination path.

```{r, eval = FALSE}
xls_url <- "http://gattonweb.uky.edu/sheather/book/docs/datasets/GreatestGivers.xls"
download.file(xls_url, here::here("data", "s008_data", "some_file.xls"), mode = "wb")
```

**NOTE: The `mode = "wb"` argument at the end is really important if you are on **
**Windows. If you omit, you will probabbly get a message about downloading a **
**corrupt file. More details about this behaviour can be found **
**[here](https://github.com/HenrikBengtsson/Wishlist-for-R/issues/30).**

Naming a file "some_file" is extremely bad practice (hard to keep track of the 
files), and you should **always** give it a more descriptive name. Often, it's 
a good idea to name the file similarly (or the same as) the original file
(sometimes that might not be a good idea if the original name is non-descriptive). 

There's handy trick to extract the filename from the URL:

```{r, eval = FALSE}
file_name <- basename(xls_url)
download.file(xls_url, here::here("data", "s008_data", file_name), mode = "wb")
```

Now we can import the file:

```{r, eval = FALSE}
read_excel(here::here("data", "s008_data", file_name))
```


## Read in a sample SPSS file. 

Let's load a sample SPSS file and work with it. Download the file from 
[here](https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/data/clevel.sav)
and save it in your `data/s008_data` folder.

These data are a random subset of the data used in [this paper](https://doi.org/10.1016/j.jvb.2018.02.005).
This was a study looking at personality traits that distinguish C-level executives
from lower-level managers among men and women. 

The subset of data here consists of 200 cases, with variables indicating:
  1. The language of assessment (English, Dutch, or French)
  2. Gender
  3. C-level or not
  4. Extraversion level, as well as 4 facet traits (Leading, Communion, Persuasive, Motivating)
  
Let's load in the data using the `haven` package.

```{r, eval = FALSE}
(clevel <- haven::read_spss(here::here("data", "s008_data", "clevel.sav")))
```

Notice that this tibble looks a little different for the `language` and `gender`
variables than normal. It has **labels** for the numeric values. This format
is what SPSS uses, but it's not standard for R. Let's convert those variables, 
and `isClevel` as well, to factors:

```{r, eval = FALSE}
clevel_cleaned <-
  clevel %>% 
  mutate(language = as_factor(language),
         gender = as_factor(gender),
         isClevel = factor(isClevel, 
                           levels = c(0, 1), 
                           labels = c("No", "Yes"))
  ) %>% 
  print()
```

Notice how the variables are now factors with labels as the entries, instead of the 
original code numbers. 


### Saving data frames

Let's save this file as a CSV file so that it's a smaller file and easier to 
import again in the future.

```{r, eval = FALSE}
write_csv(clevel_cleaned, here::here("data", "s008_data", "clevel_cleaned.csv"))
```


### Saving plots

Now let's make a plot.

```{r, eval = FALSE}
clevel_plot <-
  clevel_cleaned %>% 
  mutate(isClevel = recode(isClevel, 
                           No = "Below C-level", 
                           Yes = "C-level"),
         gender = recode(gender,
                         Female = "Women",
                         Male = "Men")) %>% 
  ggplot(aes(paste(isClevel, gender, sep = "\n"), Extraversion, color = gender)) +
  geom_boxplot() +
  geom_jitter(height = .2) +
  scale_color_manual(values = c("#1b9e77", "#7570b3")) +
  ggtitle("Extraversion Stan Scores") +
  scale_y_continuous(breaks = 1:9) +
  ggthemes::theme_fivethirtyeight() %>% 
  print()

```

Let's save the plot in several formats. This is useful if we want to use the plot
outside of Markdown. Save plots using the `ggsave()` funnction.

`ggsave()` has many options. See the help function `?ggsave` for full details.
The main arguments are `filename`, `plot`, `width` and `height` (inches by default),
and `dpi` (dots per inch; for bitmap formats).

`ggsave()` will try to guess what format you want based on the file name. If you
want, you can specify a specific format or R graphics device to save with using
the `device` argument.

```{r, eval = FALSE}
dir.create(here::here("output", "figures"))
ggsave(here::here("output", "figures", "clevel_extraversion.svg"), clevel_plot)
```

You can save to several formats. Generally, work with a _vector_ format like
`.svg`, `.eps`, or `.pdf`. Vector graphics represent the image as a series of
data points and equations. This means that they can be made smaller or larger or
zoomed in on without damaging the image quality.

If you can't use a vector format for some reason, you can also export to a _bitmap_
format. Bitmap graphs represent the image as colored dots or pixels. This means
that the image quality will suffer if you make the image larger or zoom in on it
(making it smaller can also sometimes compromise quality). With bitmap images,
you need to be concerned with *resolution* (how many pixels/dots per inch when
printed). Always use at least **300 DPI** resolution. 

There are several popular bitmap image formats. `.tiff`/`.tif` is the highest
quality, but also the largest. Use it for print graphics, but you should probably
avoid it for images to be hosted on the web. `.png` is a bit smaller, and it 
should be your go to for charts, figures, line drawings, etc. More complex 
images (e.g., photos) can get pretty big with `.png`, though. `.jpg`/`.jpeg` is
probably the most popular bitmap format. It works well for photographs hosted on
the web, but its compression often makes line drawing and charts look terrible.
`.jpg`/`.jpeg` also degrades in quality each time you edit/save it, so don't 
use it for images you intend to edit. Generally avoid `.gif` unless you are
making an animation or you need very small file size for a simple image. `.gif` 
supports very few colors, so always check your image quality after making a `.gif`. 

Let's save to some other formats:

```{r, eval = FALSE}
ggsave(here::here("output", "figures", "clevel_extraversion.eps"), clevel_plot)
ggsave(here::here("output", "figures", "clevel_extraversion.pdf"), clevel_plot)
ggsave(here::here("output", "figures", "clevel_extraversion.tiff"), clevel_plot)
ggsave(here::here("output", "figures", "clevel_extraversion.png"), clevel_plot)
```


## Organizing your project folders

Follow a consistent folder structure for all of your projects. This will make it
easier for you to say organized and make your code, data, and projects easy to 
share. 

In your project's root folder, you should have a README.md file and a .Rproj
project. Then, you should have folders that separate different types of files:

  - `data`: Stores all of your data files for a project
    - Have subfolders for different dates, waves, groups, etc. as needed
    - If you plan on saving cleaned data, having separate `data-raw` and `data`
      folders is a good idea. 
  - `markdown`: Stores your RMarkdown docuemnts. 
  - `output`: Stores any output your scripts generate
    - Depending on how many figures and other output files you will create, 
      you might want to split/subfolder this into `figures`, `reports`, etc.
    
You might add additional folders, such as:

  - `tests`: A folder that includes tests to check that your scripts or results
    are accurate. Check out the `testthat` package.
  - `templates`: A folder to hold template files (e.g., RMarkdown templates,
    Word templates, CSS files for HTML output, TeX templates for PDF output)
  - `admin`: For adminstrative documents (e.g., IRB approval, grant information)
  - `doc`: For documentation (e.g., variable codebooks, style guides), 
  - `scripts` or `src` or `R`: Folders to store functions and scripts that 
    you call from your markdown (e.g., a data import and cleaning script)
    - Use folders for each language if you are programming in multiple languages:
      `R`, `python`, `sql`, `C`, etc.
    - Can put them all in a generic `scripts` or `src` folder if not too many

## Bonus Activity: `rio`

Do the `clevel` activity again, but this time use the `rio::import()` and 
`rio::export()` functions instead of the `read_*()` and `write_*()` functions. 
In `rio::import()`, be sure to specify: `rio::import(..., setclass = "tibble")`.
