```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

# Effective Visualizations

Now that you know how to create graphics and visualizations in R, you are armed 
with powerful tools for scientific computing and analysis. But with great power 
comes great responsibility. Effective visualizations is an incredibly important 
aspect of scientific research and communication. There have been several books 
(see references) written about these principles. In class today we will be going 
through several case-studies trying to develop some expertise into making 
effective visualizations. 

## Worksheet

**The worksheet questions for today are embedded into the class notes.**

You can download this Rmd file [here](https://github.com/STAT545-UBC/Classroom/raw/master/cm013.Rmd)

Note, there will be very little coding in-class today, but I've given you plenty of exercises in the form of a supplemental worksheet (linked at the bottom of this page) to practice with after class is over.

## Resources

1. [Fundamentals of Data Visualization](https://serialmentor.com/dataviz/introduction.html) by Claus Wilke
1. [Visualization Analysis and Design](https://www-taylorfrancis-com.ezproxy.library.ubc.ca/books/9780429088902) by Tamara Munzner
1. [STAT545.com - Effective Graphics](https://stat545.com/effective-graphs.html) by Jenny Bryan
1. [ggplot2 book](https://ggplot2-book.org) by Hadley Wickam
1. [Callingbull.org](https://callingbull.org/tools.html) by Carl T. Bergstrom and Jevin West

A great resource for selecting the right plot: https://www.data-to-viz.com/.
I encourage you all to consult it when building visualizations.

## Part 1: Principles of effective visualizations

1. Apply the [principle of proportional ink](https://callingbullshit.org/tools/tools_proportional_ink.html)
    - Definition: "The amount of ink used to indicate a value should be 
      proportional to the value itself."
    - Example: Truncating the y-axis on a bar chart to exaggerate the difference 
      between bars violates the principle of proportional ink
1. Maintain a high data-to-ink ratio: [Less is more](https://speakerdeck.com/cherdarchuk/remove-to-improve-the-data-ink-ratio)
    - Definition: Remove distracting visual elements to focus attention on the data
    - Examples: Lighten line weights, remove backgrounds, never use 3D or 
      special effects, remove unnecessary/redundant labels, etc.
1. Reduce the cognitive load on viewers
    - Don't make viewers search for information or meanings
    - Examples: Use direct labels rather than legends, arrange elements in logical 
      order, don't violate expectations
1. Always update axes labels and titles on your plots
1. Choose your scale-type carefully
    - Whether you choose a linear, logarithm, sqrt scale depends on your data, 
      context, and purpose
1. Choose your graph-type carefully
    - Examples: [here](https://serialmentor.com/dataviz/directory-of-visualizations.html) is a great directory of plots
1. Choose colours with accessibility, data-fidelity, and readability in mind
    - I am colorblind! Red and black lines are identical to me.
    - Examples: [here](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette) 
      is a great set of colour schemes that are colour-blind friendly and perceptually uniform

### Make a great plot worse

Instructions: Below is a code chunk that shows an effective visualization. 
First, copy this code chunk into a new code blockk. Then, modify it to purposely 
make this chart "bad" by breaking the principles of effective visualization above. 
Your final chart still needs to run/compile and it should still produce a plot. 

```{r, message = FALSE, warning = FALSE}
library("tidyverse")

ggplot(airquality, aes(`Month`, `Temp`, group = `Month`)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(alpha = 0.3) +
    labs(x = "",
         y = "",
         title="Maximum temperature by month") +
    theme_bw() + 
    scale_x_continuous(breaks = c(5, 6, 7, 8, 9), 
                       labels = c("May", "June", "July", "August", "September")) +
    annotate("text", x = 4.08, y = 95, label="°F", size = 8) +
    coord_cartesian(xlim = c(4.5, 9.5),
                    clip = 'off') +
    theme(panel.grid.minor = element_blank(),
          panel.background = element_blank(), 
          axis.line = element_line(colour = "gray"),
          panel.border = element_blank(),
          text = element_text(size=18)
          )
```

How many of the principles did you manage to break?

## Part 2: Plot theming: The glamour of graphics

Let's watch a great video explaining and demonstrating some important principles
of the aesthetic/design choices for visualization:
[here](https://resources.rstudio.com/rstudio-conf-2020/the-glamour-of-graphics-william-chase).

You can customize the design elements of a ggplot using themes. The `theme()`
function is a general way to customize all design elements. There are also
many themes already available, such as in the `ggthemes` package.

## Part 3: Arranging plots with `patchwork`

In Assignment 3, I asked you to try to arrange two plots side by side. There is 
recently a new package called `patchwork` that makes this super easy!

```{r}
library(patchwork)
library(gridExtra)

p_mpg <- ggplot(mtcars, aes(factor(cyl), mpg, color = factor(cyl))) + 
    geom_boxplot() + theme_bw() + xlab("No. of cylinders") + ylab("MPG")

p_hp <- ggplot(mtcars, aes(factor(cyl), hp, color = factor(cyl))) + 
    geom_boxplot() + theme_bw() + xlab("No. of cylinders") + ylab("Horsepower")


```

Two other packages you might find useful for making awesome data displays:

  - `ggtext`: Lets you use (_some_) Markdown formatting in text on ggplot 
    (e.g., use images in your factor labels!)
     - italics, bold, color, superscript, subscript, font family, images
     - Install: 
         - Currently only available on GitHub: remotes::install_github("wilkelab/ggtext")
         - You _might_ also need to install the development version of `ggplot2`:
           `remotes::install_github("tidyverse/ggplot2")
     - Package docs: https://wilkelab.org/ggtext/
     - Amazing presentation:
       [here](https://resources.rstudio.com/rstudio-conf-2020/spruce-up-your-ggplot2-visualizations-with-formatted-text-claus-wilke)
       
  - `ggstatsplot`: Annotate plots with statistical test results easily
    - Install: `install.packages("ggstatsplot")`
    - Package docs: https://indrajeetpatil.github.io/ggstatsplot/
       
  - `ggpubr`: An older package for arranging plots, adding annotations and 
    data tables, etc. A bit harder to use than patchwork, but lots of code online.
    - Install: `install.packages("ggpubr")`
    - Package docs: https://rpkgs.datanovia.com/ggpubr/index.html
    - Nice tutorial: 
      [here](http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/81-ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page/)

## Supplemental worksheet: Practice scales and theming (Homework)

For homework this week, review and practice making beautiful visualizations with
ggplot. There is an extra participation worksheet you can work through to go
over the basic mechanics of formatting scales, specifying accessible colors, and 
setting theme elements in ggplots. There is also a short guided activity showing 
you how to make a ggplot interactive using plotly.

- [Supplemental worksheet here](https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/tutorials/s10_plot-theme-exercize.Rmd)