# Preparing Documents in RMarkdown

We've been using RMarkdown all semester to organize and document our analyses.
This module will focus on steps we can take to go from using RMarkdown just
as a way to run analsyes to using it to prepare polished, ready-to-submit
reports and documents.

## General formatting guidelines

There are a few things you should keep in mind when writing reports regardless
of your output format.

### Let the computer handle the formatting

Let Markdown/pandoc handle the document formatting for you. This lets you focus 
on the content as you write and will help to ensure that everything looks nice 
regardless of the format you output to. For example:

  1. Use `#`, `##`, `###` to insert headings, rather than manually entering 
     things in bold, italics, centered, etc.
  2. Use pandoc citations (e.g., `[@Wiernik2020]`) instead of manually formatting
     citations and references
     - More on this next week

### Don't manually type statistical results

This is prone to errors and is a lot of work when you need to update analyses. 
Instead let R do this for you by using output from R code chunks and inline R
code:

````markdown
`r '\x60\x60\x60{r}\nx = 5  # radius of a circle\n\x60\x60\x60'`

For a circle with the radius `r '\x60r x\x60'`, its area is `r '\x60r pi * x^2\x60'`.
````

## HTML Output

Unless you have a reason not to, use HTML. It is the most flexible format—someone
can easily read it on a computer, mobile device, table, or even print it out if
they like. You can also include many types of information displays (e.g., searchable
and filterable tables with `DT::datatable()`, movies and animations, interactive 
plots with `plotly`, etc.).

### Themes and Templates

You can change the styling of your HTML documents using themes. Specify a theme
at the top of your document in the YAML header like this:

````markdown
---
title: "Habits"
output:
  html_document:
    theme: paper
---
````

RMarkdown comes with several built-in themes. You can also design and customize
your own theme. See [here](https://bookdown.org/yihui/bookdown/theming.html)
for more details. 

Many R packages also come with templates that have specific styling. For example,
the [`tufte` package](https://bookdown.org/yihui/rmarkdown/tufte-handouts.html)
comes with templates in the styling of Edward Tufte, which look quite nice:

````markdown
---
title: "Habits"
output:
  tufte::tufte_html: default
---
````

### CSS styling

In addition to controlling your styling with themes or templates, you can also 
customize your styling further for a specific document by adding CSS code chunks. 
CSS code chunks are just like R code chunks, except that they start with `{css` 
instead of `{r`. Inside these code chunks, you can add CSS code to control the 
styling of different types of elements in your document. 

For example, let's say that you wanted to include a short caption above the code
chunks in your RMarkdown document describing the code, but without distracting
from the main text. You could do that like this:

````markdown
`r ''````{css, Style for Code Chunk Captions, echo = FALSE}
h6 {margin-bottom: 0;}
```

###### Description of the code chunk below
`r ''````{r, Chunk Label}
x = rnorm(100)
```
````

The CSS code chunk at the top adds styling for level-6 headings (written in Markdown
with 6 `######`). `margin-bottom: 0` says to not put any space between the bottom
of a level-6 heading and the next paragraph. Then, when I want to write a description
for a code chunk, I add it to the Markdown with 6 `#`. 

(Note, depending on your theme, you might need to specify additional style
information for `h6` to override heading defaults.)

For a nice introduction to styling with CSS, see [this guide](https://www.w3schools.com/Css/css_intro.asp).


### Showing and hiding code chunks

When you include code chunks in your RMarkdown documents, you can control whether
the code is shown in your rendered document using the `echo` argument:

````markdown
`r ''````{r, chunk-label, echo = TRUE}
````

If you set `echo = TRUE`, the code will be included in your rendered document, 
along with its output. If you set, `echo = FALSE`, the code will be hidden and
only the output will be shown. 

Its nice to include the code to show how the results were produced, but this can
be ugly and distracting as part of a larger document that isn't intended to
necessarily be educational. 

One of the nice things about HTML output is that you can have it both ways by
enabling the `code_folding` option in your document's YAML header at the top 
of the document. 

````markdown
---
title: "Habits"
output:
  html_document:
    code_folding: hide
---
````

`code_folding: hide` will include the code (when `echo = TRUE`), but have it
hidden by default. `code_folding: show` will show all code blocks by default, 
but still allow readers to hide them if they wish.


### Tables

So far in class, we have formatted tables using `knitr::kable()` for small 
tables and `DT::datatable()` for larger tables. Sometimes you want more control
over the formatting of your tables. There are many packages available for
customizing table formatting. See [here](https://gt.rstudio.com/#how-gt-fits-in-with-other-packages-that-generate-display-tables)
for a list of options. 

Here are a few packages that I think you should explore.


| Package | Notes |
|---------|-------|
| [`kableExtra`](https://haozhu233.github.io/kableExtra/) | Very popular. Can output to HTML or PDF, not Word (though can output to an image). |
| [`gt`](https://gt.rstudio.com/) | New package from RStudio. Designed to have a ggplot-like syntax, but for tables. Can output to HTML currently. PDF output is planned. Not yet mature. |
| [`flextable`](https://davidgohel.github.io/flextable/) | Can output to Word, PowerPoint, and HTML (and PDF sort of). |
| [`huxtable`](https://hughjonesd.github.io/huxtable/) | Can output to HTML, PDF, and Word/PowerPoint (via internal calls to the `flextable` package. |
| [`formattable`](https://renkun-ken.github.io/formattable/) | Has very useful `color_tile()` and `color_bar()` functions. Can be [combined with `kableExtra`](https://haozhu233.github.io/kableExtra/use_kableExtra_with_formattable.html). |
| `apaTables` | Formats a variety of statistical models according to APA style guidelines. |


As an example, here is a table customized using `kableExtra` to:

  1. Center the columns.
  2. Make the table the only as wide as it needs to be (instead of the full
     width of the page).
  3. Add several types of footnotes. 
  
```{r}
library(knitr)
library(kableExtra)
dt <- mtcars[1:5, 1:6]
kable(dt, align = "c") %>%
  kable_styling(full_width = FALSE) %>%
  footnote(general = "Here is a general comments of the table. ",
           number = c("Footnote 1; ", "Footnote 2; "),
           alphabet = c("Footnote A; ", "Footnote B; "),
           symbol = c("Footnote Symbol 1; ", "Footnote Symbol 2")
           )
```


## PDF Output

PDF is a format if the document is intended to be printed. It's also the format
most folks expect "final" versions of papers to be distributed in. It's not as
flexible as HTML, unfortunately.

### TeX

PDF output requires TeX, a widely-used but very fiddly, typesetting engine. The
easiest way to install and work with TeX using RMarkdown is the `tinytex` package.
Install the `tinytex` R package, then use the `tinytex::install_tinytex()` function
to install the TeX program on your computer.

```{r, eval = FALSE}
install.packages("tinytext")
tinytex::install_tinytex()
```

After you install tinytex, you can tell RMarkdown to output to PDF, either using
default `pdf_document` or `beamer_presentation` formats, or other templates, such
as `tufte::tufte_handout` or `papaja::apa6_pdf`.

````markdown
---
title: "Habits"
output:
  pdf_document:
    latex_engine: lualatex
  tufte::tufte_handout:
    latex_engine: lualatex
---
````

When you output to a PDF format, always set the `latex_engine: lualatex` argument
to get the best output and to avoid problems with mathematical and non-English
symbols.

### Tables

For PDF output, your best options for tables beyond `knitr` currenly are probably 
`kableExtra` and `huxtable`. `gt` might be a good option in the future.

### Customizing LaTeX Templates

You can control additional parameters of your PDF output (e.g., whether a table
of contents is included, formatting of various types of text or objects) by 
adding arguments to your YAML header or even supplying a custom LaTeX template.
See [here](https://bookdown.org/yihui/rmarkdown/pdf-document.html) for more details.


## Word and PowerPoint Output

RMarkdown can also output to Word or PowerPoint formats. Generally, you should 
regard these as _output_ formats, not source files—if you need to edit a document,
do it in the RMarkdown, not in the Word or PowerPoint output. 

That said, a lot of folks (even me!) find it more comfortable to write in Word
or make presentations in PowerPoint. For example, when presenting at conferences,
you might have to use PowerPoint instead of a PDF viewer. 

With that in mind, you should strive to make your Office documents as
_reproducible as possible_. For example, consider writing your Methods and 
Results sections and your Tables and Figures in RMarkdown to take advantage of 
the reproducibility of RMarkdown.  When you update these sections, update them 
in RMarkdown, render to Word, then cut-and-paste them as a whole into your 
document in the right spot. This will make updating your results fast, while 
still letting you and your collaborators work in the Word program generally. 

### Basic Word or PowerPoint output

Basic output to Word or PowerPoint is done using the built-in `word_document`
and `powerpoint_presentation` output formats. You could also use other templates,
such as `papaja::apa6_word` or `redoc::redoc` (see below).

You can provide a template Word or PowerPoint file with customized heading styles,
template text, etc. (e.g., such as [this one](https://github.com/psychmeta/psychmeta/raw/master/inst/templates/reference_docx.docx))
by specifying the `reference_docx` or `reference_pptx` arguments.

````markdown
---
title: "Habits"
output:
  word_document:
    reference_docx: word_template.docx
  powerpoint_presentation:
    reference_pptx: powerpoint_template.pptx
---
````

### Tables

For Word or PowerPoint output, your best options for tables beyond `knitr` 
currenly are probably `flextable` and `huxtable`.

### `redoc`

[`redoc`](https://noamross.github.io/redoc/) is a great package if you (or one
of your collaborators) really prefers to write in Word instead of Markdown. With 
`redoc`, you can render your RMarkdown document to Word, send it for comments 
and revisions to your colleage, then re-import the marked-up document back into 
Markdown for further editing. It's a great way to combine the familiarity of 
Word with the reproducibility of Markdown. See [here](https://noamross.github.io/redoc/) 
for a walkthrough.

### `tidystats`

The [`tidystats` package](https://www.willemsleegers.com/tidystats.html) takes
a somewhat different approach to reproducibility with Word. Instead of rendering
the whole Word document, it outputs the statistical results to a computer-readable
file Then, using a companion [Word plugin](https://github.com/WillemSleegers/tidystats-Word-add-in), 
you insert a holder for the results into your document. If you later update the 
results file by running new analyses in R, the results in Word will also update. 
It's pretty slick.

### `officedown`

If you want even more control over your Word or PowerPoint documents, the 
[`officedown`](https://github.com/davidgohel/officedown) package provides a 
variety of features to control each aspect of your file. The 
[`officer` package](https://davidgohel.github.io/officer/) provides even more 
extensive control, but doesn't use Markdown. `officer` can be useful if you output 
to Word using Markdown, but then need to make a few adjustments to the rendered 
file afterward.


## APA-style Manuscripts

If you are writing an APA-style manuscript, the [papaja package](https://crsh.github.io/papaja_man/)
can do a lot of the formatting work for you to output to Word (`papaja::apa6_word`)
or PDF (`papaja::apa6_pdf`), as well as a lot of the work to format your 
statistical results according to the APA Style Manual. Check it out!

     