## Making ggplots interactive with plotly

## Plotly demo

You can make interactive graphs and plots in R using the plotly library. 
This is a demo of what plotly is and why it's useful, and then you can try 
converting a static ggplot graph into an interactive plotly graph.

*For this demo, make sure you have the following packages installed and loaded:*

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(gapminder)
library(plotly) 
```

### Make `ggplot2` graphs interactive

It's very easy to convert an existing ggplot2 graph into an interactive graph with `plotly::ggplotly`

On the below graph, explore the interactive options:

- *Hover* your cursor over individual points
- *Zoom* in and out by dragging across / using the zoom tool
- *Single-* and *double-click* items on the legend to isolate groups of points
- While zoomed-in, use the *pan* tool to "move" around the plot, google maps style!

```{r}
p <- gapminder %>%
    ggplot(aes(x = gdpPercap, y = lifeExp, color = continent)) +
    geom_point() 

p %>%
    ggplotly()
```

### Make interactive plots with `plotly::plot_ly`

We can also make interactive graphs using the the `plotly::plot_ly` function:

```{r}
p <- gapminder %>%
    plot_ly(x = ~gdpPercap,
            y = ~lifeExp,
            color = ~continent,
            
            # mode specifies the geometric object e.g. "markers" for points, "line" for lines
            mode = 'markers',
            
            # type controls the "type" of graph e.g. 'bar', 'scatter'
            type = 'scatter'
            )

p
```

### Share with others

To share with others:

1. Create a plotly account @ [plot.ly](plot.ly)
2. Navigate to settings, and take in the following information:
- your user name
- api key 

Now, we will tell R your account information so that we can upload these plots 
to the web.

Note that once we run `api_create()`, the browser will open to a webpage 
displaying your interactive plot. You can share this page with others, but they 
will only be able to **view**. If you want others to be able to **edit** the 
graph, you can invite others to "*collaborate*" in the "*Sharing link*" option. 

```{r eval = FALSE}
# fill in the below with your information
Sys.setenv("plotly_username"="your_plotly_username")
Sys.setenv("plotly_api_key"="your_api_key")

# upload our plots to the website
api_create(p, filename = 'name-of-your-plot')
```
