## Dates and Times (5 min)

The `lubridate` package is great for identifying dates and times. You can also 
do arithmetic with dates and times with the package, but we won't be discussing 
that.

Make an object of class `"Date"` using a function that's some permutation of 
`y`, `m`, and `d` (for year, month, and date, respectively). These functions are 
more flexible than your yoga instructor:

```{r}
lubridate::mdy("September 24, 2019")
lubridate::mdy("Sep 24 2019")
lubridate::mdy("9-24-19")
lubridate::dym(c("24-2019, September", "25 2019 Sep"))
```

Notice that they display the dates all in `ymd` format, which is best for 
computing because the dates sort nicely this way.

This is not just a character!

```{r}
lubridate::ymd("2019 September 24") %>% 
  class()
```

You can tag on `hms`, too:

```{r}
lubridate::ymd_hms("2019 September 24, 23:59:59")
```

We can also extract information from these objects. Day of the week:

```{r}
today <- lubridate::ymd("2019 September 24")
lubridate::wday(today, label = TRUE)
```

Day:

```{r}
lubridate::day(today)
```

Number of days into the year:

```{r}
lubridate::yday(today)
```

Is it a leap year this year?

```{r}
lubridate::leap_year(today)
```

The newer `tsibble` package gives these `lubridate` functions some friends. 
What's the year and month? Year and week?

```{r}
tsibble::yearmonth(today)
tsibble::yearweek(today)
```


## Tsibbles (15 min)

A `tsibble` (from the package of the same name) is a special type of `tibble`, useful for handling data where a column indicates a time variable.

As an example, here are daily records of a household's electricity usage:

```{r}
energy <- here::here("data", "daily_consumption.csv") %>% 
  read_csv()
energy
```

Let's make this a `tsibble` in the same way we'd convert a data frame to a `tibble`: with the `as_tsibble()` function. The conversion requires you to specify which column contains the time index, using the `index` argument.

```{r}
(energy <- as_tsibble(energy, index = date))
```

We already see an improvement vis-a-vis the sorted dates!

This is an example of _time series_ data, because the time interval has a regular spacing. A `tsibble` cleverly determines and stores this interval. With the energy consumption data, the interval is one day ("1D" means "1 day", not "1 dimension"!):

```{r}
interval(energy)
```

Notice that there is no record for December 21, 2006, in what would be Row 5. Such records are called _implicit NA's_, because they're actually missing, but aren't explicitly shown as missing in your data set. If you don't make these explicit, you could mess up your analysis if it's anticipating your data to be equally spaced in time. Just `full_gaps()` to bring them out of hiding:


```{r}
(energy <- fill_gaps(energy))
```

Already, it's better to plot the data now that these gaps are filled in. Let's check out 2010. See how the plot without NA's can be a little misleading? Moral: always be as honest as possible with your data.

```{r}
small_energy <- filter(energy, year(date) == 2010)
cowplot::plot_grid(
  ggplot(small_energy, aes(date, intensity)) +
    geom_line() +
    theme_bw() +
    xlab("Date (in 2010)") +
    ggtitle("NA's made explicit"),
  ggplot(drop_na(small_energy), aes(date, intensity)) +
    geom_line() +
    theme_bw() +
    xlab("Date (in 2010)") +
    ggtitle("NA's in hiding (implicit)"),
  nrow = 2
)
```

How would we convert `gapminder` to a `tsibble`, since it has a time series per country? Use the `key` argument to specify the grouping:

```{r}
gapminder %>% 
  as_tsibble(index = year, key = country)
```

### `index_by()` instead of `group_by()` (5 min)

It looks like there's seasonality in intensity across the year:

```{r}
ggplot(energy, aes(yday(date), intensity)) +
  geom_point() +
  theme_bw() +
  labs(x = "Day of the Year")
```

Let's get a mean estimate of intensity on each day of the year. We'd like to `group_by(yday(date))`, but because we're grouping on the index variable, we use `index_by()` instead. 

```{r}
energy %>% 
  tsibble::index_by(day_of_year = yday(date)) %>% 
  dplyr::summarize(mean_intensity = mean(intensity, na.rm = TRUE))
```

What if we wanted to make the time series less granular? Instead of total daily consumption, how about total weekly consumption? Note the convenience function `summarize_all()` given to us by `dplyr`!

```{r}
energy %>% 
  tsibble::index_by(yearweek = yearweek(date)) %>% 
  dplyr::summarize_all(sum)
```

By the way, there's no need to worry about "truncated weeks" at the beginning and end of the year. For example, December 31, 2019 is a Tuesday, and is Week 53, but its "yearmonth" is Week 1 in 2020:

```{r}
dec31 <- "2019-12-31"
wday(dec31, label = TRUE)
week(dec31)
yearweek(dec31)
```


## Time remaining?

If so, let's make tibbles with `tibble()`, and make a list column while we're at it. Maybe even `nest()` and `unnest()`.
