# The Model-Fitting Paradigm in R

## Update Regarding Covid-19

I hope you are all doing okay and staying safe. I've posted some guidelines for 
the rest of the semester on the 
[course website](https://wiernik-datasci.netlify.com/#covid-19).

A few additional points:

  1. To be maximally flexible with people's schedules and obligations at this 
     time, there will not be live class sessions for the rest of the semester.
  2. There will not be graded homework or peer review for the rest of the semester.
  3. The only remaining graded activities will be:
     1. The final project
     2. Your participation repo 
        - Only activities before Spring Break are required; the remaining 
          activities are optional.
  4. I will post class materials tomorrow morning for you to work through on 
     your own schedule. 
  5. I am available to talk via your preferred platform (Skype, Teams, Google 
     Hangouts, FaceTime, Zoom, Phone, Smoke Signals, Carrier Pigeons) with 
     questions, feedback, etc. 
     - Please message me (text, GitHub, Twitter, email) to arrange a time.
  6. Final grading will be generous and cognizant of the current situation.


## Today

Up to now, we have been developing skills for data wrangling and 
_exploratory data analysis_ (e.g., making summary tables, visualizing trends).
Sometimes though, we would like to actually fit formal statistical models and 
use these to draw inferences and conclusions. 

Today, we are going to look at the model-fitting paradigm in R. We will see how
to fit formal models (e.g., regression models [including things like _t_ tests,
ANOVA, correlation analysis, etc.]). We will also see how to work with the
results of R model objects.

To that end, today 

1. Introduction and motivation for model-fitting in R.
1. Distinguish exploratory data analysis from model-fitting.
1. Practice a full model-fitting analysis workflow.


## Resources

- [`broom` vignette](https://cran.r-project.org/web/packages/broom/vignettes/broom.html)

- [*R for Data Science* chapter on modeling](https://r4ds.had.co.nz/model-basics.html)

If you're interested in learning more about the actual statistical/machine 
learning methods for fitting models, I highly recommend the books 
[*Learning Statistics with R*](https://learningstatisticswithr.com/) and
[*An Introduction to Statistical Learning*](https://www-bcf.usc.edu/~gareth/ISL/).
Both are freely available online and use R. *Learning Statistics with R* is
particularly well-written and funny.

```{r packages, cache = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(broom)
set.seed(1234)

theme_set(theme_minimal())
```


## What is Model-Fitting?[^thanks]

We often want to use information on one (or more) variable (the "predictor(s)") 
to learn or make judgments about another variable (the "criterion" or "response").
For example, if I know someone's height, what can I say about their weight? In
general, someone who is 7 feet tall probably weighs more than someone who is 4
feet tall.

**Example:** Consider the scatterplot below.

1. A car weighs 4000 lbs. What can we say about its mpg?
2. A car weights less than 3000 lbs. What can we say about its mpg?

```{r, fig.width=5, fig.height=3}
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(x = "Weight (1000's of lbs)")
```

In the `mtcars` data, a car's mpg and its weight are _correlated_ (they are not
"independent"). So, we could make decent predictions about its miles per gallon
if all we knew was its weight.

**Example:** What can we say about rear axle ratio if we know something about 
quarter mile time?

```{r, fig.width=5, fig.height=3}
ggplot(mtcars, aes(qsec, drat)) + 
  geom_point() +
  labs(x = "Quarter mile time",
       y = "Rear axle ratio")
```

This approach we use above relies on graphs and visual inspection to make
inferences. This graphical approach is often called _exploratory data analysis_.
(They are many methods for exploratory data analysis [EDA], but graphical methods
are some of the most common.)

Sometimes, EDA isn't enough. Perhaps we want to make more specific or precise
predictions (e.g., what is the specific range of mpg values I can expect for a
car that weights 4000 lbs?). Perhaps the relationship between two variables is
fairly weak (but not zero!), so it's hard to see a pattern just with your eyes.

In these cases, we can answer questions more precisely by _fitting a model_: 
a curve that predicts a response variable, $y$, using a predictor variable (or
variables), $x$. This is also called a __regression curve__ or a 
__machine learning model__. 

(There are more comprehensive models too, such as modeling entire distributions, 
but we will keep things simpler here.)

There are typically two goals of fitting a model:

1. Make predictions.
   - If I know a patient's diagnostic test score, how likely are they 
     attempt suicide?
   - If I know a medical school applicant's MCAT score, what GPA are they 
     expected to obtain?
2. Interpret variable relationships ("explanation").
   - Is age related to personality scores?
   - Does political orientation influence the effectiveness of different
     messaging techniques?
     
The advantage of formal modeling is that it (1) provides more specific numeric
estimates of variable relationships and predicted outcomes and (2) provides a
numeric indication of uncertainty (using uncertainty intervals).


## The Linear Model

A linear model is a specific and very useful type of model. A linear model says
that if we change the predictor variable $x$ by a certain amount, we expect the
response variable $y$ to also change by a specific (constant) amount. Most statistics
we use in psychology are linear models (e.g., correlations, t-tests, regression,
ANOVA, ANCOVA, …)

(We aren't going to dive into all of the details of linear models [or other types
of models] or their assumptions. You can learn more about those in your stats
classes or using the books in the **Resources** section. For now, focus on the 
basic idea of the linear model as quantifying the [linear] relationship between
variables and estimating numeric values to accompany your plots.)

Linear models have the generic form

$$y = \beta_0 + \beta_1 \times x$$

where $y$ is the **outcome of interest**, $x$ is the **explanatory** or **predictor**
variable, and $\beta_0$ and $\beta_1$ are **parameters** that reflecct the relationship
between the two variables (i.e., _"how do I use information on $x$ to predict a 
value on $y$?"). $\beta_1$ is the _slope_ (how much does $y$ change for each change
of 1 in $x$?). $\beta_0$ is the _intercept_ (what is the expected value of $y$ 
when $x$ is _0_?). 

We use the observed data for $x$ and $y$ to generate a **fitted model**, where 
we make our best guess as to the true values for the $\beta_0$ and $\beta_1$ 
parameters by picking values that best fit the data.

Let's look at some simulated data with predictor $x$ and response $y$. 

```{r sim-plot}
ggplot(modelr::sim1, aes(x, y)) + 
  geom_point()
```

This looks like a linear relationship. We could randomly draw prediction lines. 
(That is, we could randomly generate values for $\beta_0$ and $\beta_1$ for the 
formula $y = \beta_0 + \beta_1 \times x$ to try and explain or predict the 
relationship between $x$ and $y$.):

```{r sim-random-fit}
models <- tibble(
  a1 = runif(250, -20, 40),
  a2 = runif(250, -5, 5)
)

ggplot(modelr::sim1, aes(x, y)) + 
  geom_abline(aes(intercept = a1, slope = a2), data = models, alpha = 1/4) +
  geom_point()
```

Obviously some of these lines are better than others (some values of $\beta_0$ 
and $\beta_1$ better capture the true relationship between $x$ and $y$). 

We need a definition of "better" to separate good sets of parameter values from 
bad sets of parameter values. One approach that is widely used called **least squares**.
Least squares means that $\beta_0$ and $\beta_1$ are chosen to _minimize_ the 
_sum of the squares of the errors_ of the predictions made by the model. That is,
which line can we draw that will get as many of the data points as close as possible
to that line? The errors in prediction (how far are the points from their predicted
values [the line]?) are the _vertical_ difference between the actual values for 
$y$ and the predicted values for $y$ (the points on the line).

```{r sim-error, echo = FALSE}
dist1 <- modelr::sim1 |> 
  mutate(
    dodge = rep(c(-1, 0, 1) / 20, 10),
    x1 = x + dodge,
    pred = 7 + x1 * 1.5
  )

ggplot(dist1, aes(x1, y)) + 
  geom_abline(intercept = 7, slope = 1.5, color = "grey40") +
  geom_point(color = "grey40") +
  geom_linerange(aes(ymin = y, ymax = pred), color = "#3366FF")
```

[R for Data Science](http://r4ds.had.co.nz/model-basics.html) gives a good 
introduction to performing these calculations manually by writing your own 
functions. I encourage you to read through and practice some of this code, 
especially if you have no experience with linear models.

However for our purposes, let's focus on using existing R tools to fit a linear
model. You can use `ggplot2` to draw the best-fit line:

```{r sim-plot-lm}
ggplot(modelr::sim1, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm")
```

The line in blue is the best-fit line. The gray band is the 95% confidence/compatibility 
intervals for the predicted values. A confidence interval indicates the range of 
true parameter values that are reasonably compatible with your data (1) assuming 
your model assumptions are correct (e.g., the relationship between the variables 
is linear) and (2) with a specified Type 1 error rate (e.g., a 95% confidence 
interval is expected to miss the true parameter value 5% of the time). (For more 
discussion on the meaning and interpretation of confidence/compatibility, see
[this discussion](https://doi.org/10.1136/bmj.l5381).)

The `method = "lm"` part of `geom_smooth()` specifies what type of model to use
to fit the line. `"lm"` stands for "linear model" and uses R's function for 
fitting linear models, `lm()`, to fit the line. `lm()` is one of many modeling 
function in R. 

## Model-Fitting in R

Model fitting methods usually use a common format in R:

```
method(formula, data, options)
```

They also tend to have a common output object: a special _list_. 

__method__:

The modeling function, such as:

- Linear Regression: `lm()`
- Generalized Linear Regression: `glm()` (e.g., logistic regression, Poisson regression)
- Local regression (aka "smoother lines"): `loess()`
- Quantile regression: `quantreg::rq()`
- Meta-analysis: `metafor::rma()`
- ...

For now, we are only goint to use `lm()`.

__formula__:

Formulas are used to describe the structure of the model you want to fit. What
is the response variable you are predicting? What are the predictor variables?

In R, formulas take the form `y ~ x1 + x2 + ... + xp`.

The left side of the formula (before `~`) is the response variable. The right side
of the formula (after `~`) are the predictors. Separate multiple predictors using
`+`. (We will get to some more complex formulas later down the page.)

`y`, `x1`, `x2`, etc. are the column names in your data frame. R will look into 
your data frame and look for the columns with the names given in your formula
(much like `aes()` in ggplot).

__data__: The data frame.

__options__: Additional arguments specific to the method and function.


### Example model using `gapminder`:

1. Fit a linear regression model to life expectancy ("Y") from year ("X") by 
   filling in the formula. 
   
First, create a subset of the `gapminder` dataset containing only data from Europe

```{r}
(gapminder_europe <- filter(gapminder, continent == "Europe"))
```

Now, use the `lm()` function to fit a linear model:

```{r}
(mod_europe <- lm(lifeExp ~ year, data = gapminder_europe))
```

When we print `mod_europe`, it shows us the "coefficients" for the model (the
estimated values of $\beta_0$ and $\beta_1$). Does that mean that the life 
expectency at "year 0" was equal to -397.7646?!

No, that doesn't make much sense. We need to be careful to only interpret the
predictions of the model for the general range of the variables we actually
have in our data. 

We can modify how we fit the model to make the value of the `(Intercept)` ($\beta_0$)
more meaningful. Let's fit it so that it reflects the predicted life expectancy 
value for the first year in the data, 1952. To do that, we use the `I()` function. 
`I()` lets you tell R to fit the model to an arithmetic transformation of a 
variable. For example:

```{r}
(mod_europe <- lm(lifeExp ~ I(year - 1952), data = gapminder_europe))
```

In this model, the `(Intercept)` coefficient ($\beta_0$) is the predicticed life
expectancy in 1952, and the `I(year - 1952)` coefficient ($\beta_1$) is the expected
increase in life expectancy each year.

What type of R object is the output of `lm()`? Let's see.

```{r}
class(mod_europe)
is.list(mod_europe)
```

`mod_europe` is an object of class `"lm"` (an `lm` model object). This is a special
type of `list`. A `list` is a vector than can contain all manner of different types
of values, and the values don't have to all be the same or be the same length.

Let's look at the structure of an `"lm"` model object:

```{r}
str(mod_europe)
```

An `"lm"` model object contains a lot of important information, some of which 
you may recognize but most of it you do not. It's a bit of a mess inside (don't 
tell its mom). It can be easier to see what slots an object like this contains 
using the `names()` function:

```{r}
names(mod_europe)
```

We can view a summary of the model results using the `summary()` function:

```{r}
summary(mod_europe)
```

Under the hood, the result of `summary(mod_europe)` is another list that contains
even more information:

```{r}
names(summary(mod_europe))
```

As you can see, model objects are decidely _*not*_ tidy. So, we aren't going to 
pull results directly from them. Instead, we are going to use functions from base
R and the `broom` package to tidy up and navigate model results.

### Predicted Values

One basic thing we might want to do is extract the predicted (or "fitted") values
for the model. We can do this using the `predict()` function:

```{r}
predict(mod_europe)
```

`predict(mod_europe)` alone returns just the fitted values. We can additionally
get a confidence interval for these fitted values by adding `interval = "conf"`:

```{r}
predict(mod_europe, interval = "confidence") # can abbreviate to "conf"
```

Or we can predict on a new dataset:

```{r}
new_data_france <- tibble(year = c(1955, 1964, 1971, 1998, 2000, 2006))
predict(mod_europe, newdata = new_data_france, interval = TRUE)
```

(If you want a _prediction interval_ [what is the whole predicted _range_ of 
values, not just the predicted mean value, we can expect in new data], 
specify `interval = "prediction"` or `interval = "pred"` instead.)

### Residuals

We might also want to extract the model *residuals*; the differences between
the model predicted values and the actual data for each case. Residuals are useful
for evaluating model fit and other diagnostics.

```{r}
resid(mod_europe)
```

### `broom::augment`

The `augment()` function from the `broom` package can compute fitted and residual
values, as well as several model diagnostic statistics for each data point.
`augment()` always returns a data frame, so it is more convenient to use for things
like plotting than `predict` or `resid`. It doesn't currently return 
confidence/prediction intervals for `lm` objects, though.

```{r}
augment(mod_europe, data = gapminder_europe)
```

A linear model assumes that the residuals are normally distributed, so looking at
a plot of the residuals is useful to determine if our model is poorly fit:

```{r}
augment(mod_europe, data = gapminder_europe) |> 
  ggplot(aes(x = .resid)) +
  geom_histogram() +
  theme_minimal() +
  xlab("Model residuals") +
  ylab("Count")
```

A plot of the residuals against the observed or fitted values can also help to 
show if a specific range of values is more poorly predicted than others:

```{r}
augment(mod_europe, data = gapminder_europe) |> 
  ggplot(aes(x = year, y = .resid)) +
  geom_point() +
  theme_minimal() +
  xlab("Year") +
  ylab("Model residuals")
```


### Model coefficients

If we are interested in using our model to understand relationships between the
variables, we will also want to examine and interpret the model coefficients
$\beta_0$ and $\beta_1$ themselves. 

We can view the model coefficients using the `coef()` function:

```{r}
coef(mod_europe)
```

We can get more information about the coefficients, such standard errors and
hypothesis tests using `summary()`:

```{r}
summary(mod_europe)
```

We can get confidence intervals for the coefficients using the `confint()` function:

```{r}
confint(mod_europe)
```

Rather than using all of these indivdual functions, the `broom::tidy()` function
puts them all together and conveniently returns the results as a data frame
(great for tabling or plotting):

```{r}
tidy(mod_europe, conf.int = TRUE)
```

Looking at these results, we can see that the predicted average life expectancy
for European countries in 1952 is `r  tidy(mod_europe, conf.int = TRUE)[1, "estimate"]` years
[with a 95% confidence interval of `r  tidy(mod_europe, conf.int = TRUE)[1, c("conf.low", "conf.high")]`, indicating a range of true values that are reasonably compatible with our
data]. Each year, the predicted life expectancy increases by `r  tidy(mod_europe, conf.int = TRUE)[2, "estimate"]` years [95% CI `r  tidy(mod_europe, conf.int = TRUE)[1, c("conf.low", "conf.high")]`].


### Model fit

Finally, we might want to evaluate overall how well our model fits the data. For
example, how well does `year` alone account for all of the variability in life
expectancy in European countries over time? We can obtain a variety of model fit
statistics using the `broom::glance()` function.

```{r}
glance(mod_europe)
```

In this data frame, `adj.r.squared` is the squared correlation between the 
predicted values and the actual values for life expectancy. The square root of this
value is useful for evaluating model fit. 
Here, _R_ = `r sqrt(glance(mod_europe)$adj.r.squared)`, indicating a very strong 
relationship between year and life expectancy, but there is still some variability
left over across countries in a single year. The `sigma` value is the left over
(residual) standard deviation of the response variable (life expectancy) after
accounting for the predictors (year). Here, within a single year, countries'
life expectancies have a standard deviation of `r glance(mod_europe)$sigma`.

If you want a full ANOVA table for your model, use the `anova()` function:

```{r}
anova(mod_europe)
```

(I prefer to focus on the adjusted $R^2$ and `sigma` values to the ANOVA table.)

## Categorical and Multivariable Models

If you include a character or factor variable as a predictor, R will turn this
into a series of dummy-coded contrast variables:

```{r}
(mod_europe_country <- lm(lifeExp ~ country, data = gapminder_europe))
```

You can fit models with multiple predictors by adding them to the right side of
your formula:

```{r}
(mod_europe_gdp <- lm(lifeExp ~ year + gdpPercap, data = gapminder_europe))
```

You can transform variables before including them in the model:

```{r}
(mod_europe_lgdp <- lm(lifeExp ~ year + log(gdpPercap), data = gapminder_europe))
```

You can specify squared terms or other arithmetic transformations using the 
`I()` function:

```{r}
(mod_europe_yrsq <- lm(lifeExp ~ year + I(year^2), data = gapminder_europe))
```

You can specify interactions between two variables using `*`. This will include
both variables themselves (`year`, `log(gdpPercap)`) and their product/interaction
(`year:gdpPercap`):

```{r}
(mod_europe_interaction <- lm(lifeExp ~ year * log(gdpPercap), data = gapminder_europe))
```

You can use `anova()` to compare two nested models:

```{r}
anova(mod_europe, mod_europe_yrsq)
```

This comparison suggests that adding a squared term for year only slightly
improves model fit. The year-trend for life expectancy is most linear. 


## Summary

Model-fitting is about making specific numeric predictions about 
(1) individual scores or (2) relationships between variables. It is important
to evaluate models using both (1) the estimated/predicted values for individual
data points and model coefficients, as well as (2) an uncertainty interval that
indicates a range of plausible true values that are compatible with your data.

After fitting a model using `lm()` or another modeling function, you can use
`summary()`, `predict()`, `resid()`, `confint()`, and other functions to explore
model results. The `broom::augment()`, `broom::tidy()`, and `broom::glance()`
packages are particularly useful for extracting a variety of useful statistics
in an easy-to-use format. You can then pass these results to `knitr::kable()`,
`ggplot()`, or other functions to report and present them.


## Activity

Use the `psych::bfi` dataset. Compute mean scores for each of the Big Five scales
for each person. Then, fit linear models to answer the following questions. Present
your results using both tables of results and figures.

1. Do men and women differ on the Big Five traits? How big are the differences?
2. Do the Big Five traits increase or decrease with Age? Is there a linear or 
   squared trend?
3. Do the Big Five traits differ across educational levels? Treat education as 
   a categorical variable.
4. How well do age and gender together predict the Big Five traits?
5. In your models in part (4), do the residuals appear to be normally distributed?
   Are they consistent across age ranges and gender groups?


[^thanks]: Some parts of this tutorial are adapted from a
[tutorial](https://cfss.uchicago.edu/notes/linear-models/) by Dr. Benjamin Soltoff.
