- Making tibbles using `tibble::tibble()` and `tidyr::expand_grid()`.

## Making tibbles

In base R, we can make data frames using the `data.frame()` function. The tidyverse version is `tibble::tibble()`, which also has backwards referencing to variables you make on the fly. It's also stricter by not allowing recycling unless the vector is of length 1:

Good:

```{r}
tibble(x = 1:6,
       y = min(x))
```

Bad:

```{r, error = TRUE}
tibble(x = 1:6,
       y = 1:2)
```

Truly manual construction of tibbles is easy with `tibble::tribble()`:

```{r}
tribble(
  ~Day, ~Breakfast,
  1, "Apple",
  2, "Yogurt",
  3, "Yogurt"
)
```

Check out the `datapasta` package for ways to reproducibly copy-paste data from 
spreadsheets into R. `datapasta` uses `tribble()`.

List columns are easy with tibbles!

```{r}
(list_col <- tibble(n = 1:2,
                    y = list(iris, mtcars)))
```

Often obtained with `nest()` and `unnest()`:

```{r}
(iris_nest <- iris |> 
  group_by(Species) |> 
  nest())
```

```{r}
iris_nest |> 
  unnest(data)
```

`expand_grid()` to obtain all combinations:

```{r}
expand_grid(x = 1:2, y = 1:2, z = 1:2)
```

In conjunction with `nesting()`:

```{r}
expand_grid(nesting(x = 1:2, y = 1:2), z = 1:2)
```


## Implicit `NA`'s

Sometimes there's "hidden" missing data in a tibble. Here's an example  
from the documentation of `tidyr::expand()`:

```{r}
(df <- tibble(
  year   = c(2010, 2010, 2010, 2010, 2012, 2012, 2012),
  qtr    = c(   1,    2,    3,    4,    1,    2,    3),
  return = rnorm(7)
))
```

Here, there are values for `qtr` 1, 2, 3, 4 in `year` 2010 and 
`qtr` 1, 2, 3 in `year` 2012, but no `qtr` 4 in `year` 2012. 
The value of `return` in `year` 2012, `qtr` 4 is "implicilty" `NA`.

We can use functions to fill out this data frame and make these "implicit" missing
values explicit.

We can consider all existing combinations by invoking the column names in 
`expand()` or `complete()` (which either _drops_ or _keeps_ all other columns):

```{r}
df |> 
  expand(year, qtr)
df |> 
  complete(year, qtr)
```

We can consider new combinations by specifying an expectation of possible values:

```{r}
df |> 
  expand(year = 2010:2012, qtr)
df |> 
  complete(year = 2010:2012, qtr)
```

Want to link two or more columns when looking for combinations? Use `nesting()`.


## Activity (10 min)

Fill out __Exercise 3: Making tibbles__ in the 
[worksheet](https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/tutorials/s06_pivot-exercise.Rmd).


## Exercise 3: Making tibbles

1. Create a tibble that has the following columns:

- A `label` column with `"Sample A"` in its entries.
- 100 random observations drawn from the N(0,1) distribution in the column `x`
  - "N" means the normal distribution. "(0, 1)" means mean = 0, sd = 1.
  - Use `rnorm()`
- `y` calculated as the `x` values + N(0,1) error. 

```{r}
n <- 100
FILL_THIS_IN(label = FILL_THIS_IN,
             FILL_THIS_IN = rnorm(n),
             FILL_THIS_IN)
```


2. Generate a Normal sample of size 100 for each combination of the following 
means (`mu`) and standard deviations (`sd`).

```{r}
n <- 100
mu <- c(-5, 0, 5)
sd <- c(1, 3, 10)
FILL_THIS_IN(mu = mu, sd = sd) |> 
  group_by_all() |> 
  mutate(z = list(rnorm(n, mu, sd))) |> 
  FILL_THIS_IN
```


3. Fix the `experiment` tibble below (originally defined in the documentation 
of the `tidyr::expand()` function) so that all three repeats are displayed for 
each person, and the measurements are kept. Some code is given, but it doesn't
quite work. It needs a few adjustments. What are they?

```{r}
experiment <- tibble(
  name = rep(c("Alex", "Robert", "Sam"), c(3, 2, 1)),
  trt  = rep(c("a", "b", "a"), c(3, 2, 1)),
  rep = c(1, 2, 3, 1, 2, 1),
  measurement_1 = runif(6),
  measurement_2 = runif(6)
)
experiment |> expand(name, trt, rep)
```
