---
title: "Worksheet: The Model-Fitting Paradigm in R"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(gapminder)
library(broom)
```

So you want to fit a model to your data. How can you achieve this with R?

Topics:

1. What _is_ model-fitting?
2. How do we fit a model in R?
3. How can we obtain tidy results from the model output?





## Fitting a model in R





We can plot models (with one predictor/_X_ variable) using `ggplot2` using the
`geom_smooth()` function. Specifying `method = "lm"` gives us the linear 
regression fit (but only visually!):

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    geom_point() +
    geom_smooth(method="lm") +
    scale_x_log10()
```

Let's consider another country "Zimbabwe", which has a unique behavior in the 
`lifeExp` and `year` relationship.

```{r}
gapminder_Zimbabwe <- FILL_THIS_IN
gapminder_Zimbabwe |> ggplot(aes(year, lifeExp)) + geom_point()
```

Let's try fitting a linear model to this relationship
```{r}
ggplot(gapminder_Zimbabwe, aes(year,lifeExp)) + geom_point()+geom_smooth(method = "lm", se = F)
```

Now we will try to fit a second degree polynomial and see what would that look like.

```{r}
ggplot(FILL_THIS_IN) + geom_point() + geom_smooth(FILL_THIS_IN)

```

```{r}
lm_linear <- lm(data = gapminder,formula = FILL_THIS_IN)
lm_poly <- lm(data = gapminder,formula = FILL_THIS_IN))
```


## Regression with categorical variables

```{r}
(lm_france_cat <- lm(gdpPercap ~ I(year - 1952) + continent, data = gapminder))

```

How did R know that continent was a categorical variable?
```{r}
class(gapminder$continent)
levels(gapminder$continent)
contrasts(gapminder$continent)
```

How can we change the reference level?
```{r}
gapminder$continent <- relevel(gapminder$continent, ref = "Oceania")
```

Let's build a new model
```{r}
lm_cat2 <- lm(gdpPercap ~ I(year - 1952) + continent, data = gapminder)

```


## Broom

Let's make it easier to extract info, using the `broom` package. There are three crown functions in this package, all of which input a fitted model, and outputs a tidy data frame.

1. `tidy`: extract statistical summaries about each component of the model.
    - Useful for _interpretation_ task.
2. `augment`: add columns to the original data frame, giving information corresponding to each row.
    - Useful for _prediction_ task.
3. `glance`: extract statistical summaries about the model as a whole (1-row tibble).
    - Useful for checking goodness of fit.

Exercise: apply all three functions to our fitted model, `my_lm`. What do you see?

```{r}

```
