---
title: 's04: `dplyr` Exercise'
output: 
  html_document:
    theme: paper
---

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

# Intro to `dplyr` syntax

Load the `gapminder` and `tidyverse` packages.
    - This loads `dplyr`, too.

```{r load packages}
# load your packages here:
library(FILL_THIS_IN)
library(FILL_THIS_IN)
```
    

## `select()`

1. Make a data frame containing the columns `year`, `lifeExp`, `country` from 
the `gapminder` data, in that order.

```{r}
select(gapminder, FILL_THIS_IN)
```

2. Select all variables, from `country` to `lifeExp`.

```{r}
# This will work:
select(gapminder, country, continent, year, lifeExp)

# Do it shorter with `:` :
select(gapminder, FILL_THIS_IN)
```

3. Select all variables, except `lifeExp`.

```{r}
select(gapminder, FILL_THIS_IN)
```

4. Put `continent` first. Hint: use the `everything()` function.

```{r}
select(gapminder, FILL_THIS_IN, FILL_THIS_IN)
```

5. Rename `continent` to `cont`.

```{r}
# compare
select(gapminder, FILL_THIS_IN)

rename(gapminder, FILL_THIS_IN)
```


## `arrange()`

1. Order by year.

```{r}
arrange(gapminder, FILL_THIS_IN)
```

2. Order by year, in descending order.

```{r}
arrange(gapminder, FILL_THIS_IN)
```

3. Order by year, then by life expectancy.

```{r}
arrange(gapminder, FILL_THIS_IN, FILL_THIS_IN)
```


## Piping, `%>%`

Remember: Think of `%>%` as the word "then"!

Do `select()` Task 1 and `arrange()` Task 3 together with a pipe. 

Select `year`, `lifeExp`, `country` from the `gapminder`, then arrange in order by year and life expectancy.

```{r}

```


# Transforming datasets

## `filter()`

1. Only keep data with population greater than 100 million.

```{r}
gapminder %>%
  filter(FILL_THIS_IN)
```

2. Only keep data _from Asia_ with population greater than 100 million.

```{r}
gapminder %>%
  filter(FILL_THIS_IN)
```

3. Only keep data _from the countries Brazil and China_ with population greater than 100 million.

```{r}
gapminder %>%
  filter(FILL_THIS_IN)
```

## `mutate()`

The `mutate()` function _creates_ new columns in the tibble by transforming other variables. 
Like `select()`, `filter()`, and `arrange()`, the `mutate()` function also takes a tibble as its first argument, and returns a tibble. 

The general syntax is:

```
mutate(tibble, NEW_COLUMN_NAME = CALCULATION)
```

Let's get: 

- `grossGDP`: Multiplying GPD per capita by population
- `gdpBill`: GDP in billions, rounded to two decimals.

```{r}
gapminder %>%
  mutate(FILL_THIS_IN)
```

Notice the backwards compatibility! No need for loops!

Try the same thing, but with `transmute` (drops all other variables). 

```{r}
gapminder %>%
  transmute(FILL_THIS_IN)
```

The `if_else` function is useful for changing certain elements in a data frame.

Example: Suppose Canada's 1952 life expectancy was mistakenly entered as 68.8 in the data frame, but is actually 70. Fix it using `if_else` and `mutate`. 

```{r}
gapminder %>%
  mutate(FILL_THIS_IN)
```

Your turn: 
Make a new column called `cc` that pastes the country name followed by the continent, separated by a comma.
(Hint: use the `paste()` function with the `sep=", "` argument).

```{r}
```

These functions we've seen are called __vectorized functions__—they are designed to work with vectors and do their computation on each element of the vector(s).


# Back to the Big Group

Let's head back to the main group. 


# `across()`

The `across()` function makes it easier to do `mutate()` and `summarize()` on multiple columns at a time.
When you use `across()`, you give either:
(1) A **vector** of column names, or 
(2) A **condition** that columns must match.
Then, `mutate()` or `summarize()` will be performed for each matching column.

Let's use `across()` to rescale the personality scores in `psych::bfi` to be on a 0-100 scale instead of a 1-6 scale.

```{r}
psych::bfi %>%
  as_tibble(rownames = ".id") %>% 
  mutate(across(A1:O5, ~ (.x - 1) / 5 * 100))
```

Now, let's use `across()` to **mean center** all of the numeric variables in gapminder:

```{r}
gapminder %>% 
  filter(country == "Canada")

gapminder %>% 
  filter(country == "Canada") %>% 
  mutate(across(where(is.numeric), ~ .x - mean(.x, na.rm = TRUE)))
```

You can check out other ways to select matching variables by running:
`?tidyselect::select_helpers`.

Now you try.
Use `psych::bfi` and mean center the personality scales. 

```{r}
psych::bfi %>%
  as_tibble(rownames = ".id") %>% 
  FILL_THIS_IN(FILL_THIS_IN)
```


# `group_by()` and `summarize()`

`summarize()` is a way to **reduce** a data frame to summary values. 
Like `mutate()`, `summarize()` will compute a new column, but each column needs to reduce to a **single value**.

```{r}
gapminder %>% 
  summarize(mu = mean(lifeExp),
            sigma = sd(lifeExp))
```

This is most useful if combined with `group_by()`.
When you group a data frame, all operations on that data frame are done **separately for each group**. 

```{r}
gapminder %>% 
  group_by(year)
```

See how the printing of the tibble now says it is grouped?

```{r}
gapminder %>% 
  group_by(year) %>% 
  summarize(mu = mean(lifeExp),
            sigma = sd(lifeExp))
```

You can group by multiple variables at a time:

```{r}
gapminder %>% 
  group_by(year, continent) %>% 
  summarize(mu = mean(lifeExp),
            sigma = sd(lifeExp))
```

You can also combine `group_by()` with `mutate()` to create new variables by group:

```{r}
gapminder %>% 
  group_by(year) %>% 
  mutate(lifeExp_center = lifeExp - mean(lifeExp, na.rm = TRUE)) %>% 
  ungroup()
```

When you use `group_by()` with `mutate()`, always `ungroup()` at the end so that future operations happen on the whole dataset rather than by group!



# Exercises for grouped data frames

Let's do some practice with grouping (and ungrouping) and summarizing data frames!

1. 
(a) What's the minimum life expectancy for each continent and each year? 
(b) Add the corresponding country to the tibble, too. 
(c) Arrange by min life expectancy.

```{r}
gapminder %>% 
  group_by(FILL_THIS_IN) %>% 
  FILL_THIS_IN(min_life = min(lifeExp))
```


2. 
Let's compute the mean Agreeableness score across items for each participant in the `psych::bfi` dataset. 
Be sure to handle `NA`!

```{r}
psych::bfi %>%
  as_tibble(rownames = ".id") %>% 
  select(.id, A1:A5) %>% 
  rowwise() %>% 
  mutate(A_mean = mean(c(A1, A2, A3, A4, A5), na.rm = TRUE)) %>% 
  ungroup()
```

You can do this more compactly with the `c_across()` function.
this function selects columns the same way across does `across()`, then puts the matching values together with `c()`.

```{r}
psych::bfi %>%
  as_tibble(rownames = ".id") %>% 
  select(.id, education, A1:A5, O1:O5) %>% 
  rowwise() %>% 
  mutate(A_mean = mean(c_across(A1:A5), na.rm = TRUE),
         O_mean = mean(c_across(starts_with("O"), na.rm = TRUE))) %>% 
  ungroup()
```

Now compute mean scores for the other Big Five traits, as well as 
`sd` and `min` scores for each person.

```{r}

```

3.
Use your data from Question 2 to compute the **mean** score for each Big Five trait **for each education level**

```{r}

```


4. 
Calculate the growth in population since the **first year on record** 
_for each country_.
Do it by rearranging the following lines, and filling in the 
`FILL_THIS_IN`. 
To get the first value in a vecotr, use `dplyr::first()`. 

```
mutate(rel_growth = FILL_THIS_IN) %>% 
arrange(FILL_THIS_IN) %>% 
gapminder %>% 
knitr::kable()
group_by(country) %>% 
```

```{r}

```


5. 
Determine the country on each continent, that had the **largest increase in life expectancy** between two measurements.
Sort the output by the size of the increase.
To compute the the difference between rows of data, try `dplyr::lag()`.

```
gapminder %>% 
filter(lifeExp_incr == max(lifeExp_incr, na.rm = TRUE)) %>% 
group_by(country) %>% 
group_by(continent) %>% 
ungroup() %>% 
ungroup() %>% 
arrange(lifeExp_incr)
mutate(lifeExp_incr = lifeExp - lag(lifeExp)) %>% 
```

```{r}

```


# Bonus Exercises

If there's time remaining, we'll practice with these exercises. 

1. In `gapminder`, take all countries in Europe that have a GDP per capita 
   greater than 10000, and select all variables except `gdpPercap`. 
   (Hint: use `-`).

2. Take the first three columns of `gapminder` and extract the names.

3. In `gapminder`, convert the population to a number in billions.

4. Take the `palmerpenguins::penguins` data frame and extract all columns that start with the word "bill". 
    - Hint: take a look at the "Select helpers" documentation by running the 
      following code: `?tidyselect::select_helpers`.

5. Filter the rows of `palmerpenguins::penguins` for bill_lenth_mm >= 40 and flipper_length_mm >= 200.
