---
title: "ggplot2 exercises: Fix the plots"
output: 
  html_document:
    theme: paper
---

In this worksheet, we'll be looking at some problematic plots and fixing them. 

I think you might not have these two packages installed:

```
install.packages("ggridges")
install.packages("scales")
```

The exercises below include some data wrangling function. It's okay if you
aren't familiar with them all yet! We will get into a lot of them over the
next few weeks, but see if you can figure out what they are doing as you go.

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(gapminder)
library(ggridges)
library(scales)
```

<!-- The following chunk allows errors when knitting -->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```


## Exercise 1: Overlapping Points

Fix the overlapping data points problem in the following plot using each of these solutions:
  
  1. Adding an `alpha` argument
  1. Adding a `size` argument
  1. Jittering the points

```{r, fig.width = 5, fig.height = 2}
ggplot(mpg) +
  aes(cty, hwy) +
  geom_point()
```


## Exercise 2: Line for each Country

Fix this plot so that it shows life expectancy over time with a separate line
_for each country_. 
ggplot(gapminder) +

```{r, fig.width = 5, fig.height = 2}
  aes(year, lifeExp) +
  geom_line()
```


## Exercise 3: More gdpPercap vs lifeExp

### 3(a) Facets

- Change the x-axis text to be in "comma format" with `scales::comma_format()`.
- Separate each continent into sub-panels.

```{r, fig.width = 5, fig.height = 2}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp) +
  geom_point(alpha = 0.2) +
  scale_x_log10()
```

### 3(b) Bubble Plot

- Put the plots in one row, and free up the axes (let them have different scales).
- Make a bubble plot by making the size of the points proportional to population. 
  - Try adding a `scale_size_area()` layer too
- Use `shape=21` to distinguish between `fill` (interior) and `color` (exterior). 

```{r, fig.width = 5, fig.height = 2}
gapminder_noOceania <- filter(gapminder, continent != "Oceania")

ggplot(gapminder_noOceania) +
  aes(gdpPercap, lifeExp) +
  facet_wrap(~ continent) +
  geom_point(alpha = 0.2) +
  scale_x_log10(labels = scales::comma_format())
```

A list of shapes can be found [at the bottom of the `scale_shape` documentation](https://ggplot2.tidyverse.org/reference/scale_shape.html).

### 3(c) Size "not working"

Instead of alpha transparency, suppose you're wanting to fix the overplotting issue by plotting small points. Why is this not working? Fix it.

```{r, fig.width = 5, fig.height = 2}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, size = 0.1) +
  geom_point() +
  scale_x_log10(labels = scales::dollar_format())
```


## Exercise 4: Walking caribou

The following mock data set marks the (x,y) position of a caribou at four time points. 

- Fix the plot below so that it shows the **path** of the caribou.
- Add an arrow with `arrow = arrow()`.
- Add the `time` label with `geom_text()`.

```{r, fig.width = 5, fig.height = 2}
caribou_path <- 
  tribble(
    ~time, ~x, ~y,
    1, 0.3, 0.3,
    2, 0.8, 0.7,
    3, 0.5, 0.9,
    4, 0.4, 0.5
  )

ggplot(caribou_path) +
  aes(x, y) + 
  geom_line()
```

## Exercise 5: Life expectancies in Africa

### 5(a) Unhiding the data

Fix the plot so that you can actually see the data points. 

There is also the problem of overlapping text in the x-axis labels. 
How could we solve that?
  - Try to add `guide = guide_axis(n.dodge=3)` to the x scale

```{r, fig.width = 5, fig.height = 2}
americas <- filter(gapminder, continent == "Americas")

ggplot(americas) + 
  aes(country, lifeExp) +
  geom_point() +
  geom_boxplot() +
  scale_x_discrete()
```

### 5(b) Ridgeplots

We're starting with the same plot as above, but instead of the points + boxplot, try a ridge plot instead using `ggridges::geom_density_ridges()`, and adjust the `bandwidth`.

```{r, fig.width = 5, fig.height = 2}
ggplot(americas) +
  aes(country, lifeExp) +
  geom_point() +
  geom_boxplot()
```

## Exercise 6: Bar plot madness

### 6(a) Colour and stacking madness

- Change the following plot so that it shows _proportion_ on the y-axis, not count.
- Change the x-axis so that it doesn't appear to be continuous.
  - Hint: Transform the variable!
- Also change the colors of the bar fills, as well as the lines.
- Put the bars for transmission side-by-side with their own colour.
- Capitalize the legend title.

```{r, fig.width = 5, fig.height = 2}
mtcars2 <- mutate(mtcars, transmission = if_else(am == 0, "automatic", "manual"))
ggplot(mtcars2) +
  aes(cyl) +
  geom_bar(aes(color = transmission))
```

### 6(b) Bar heights already calculated

Here's the number of people having a certain hair colour from a sample of 592 people:

```{r}
(hair <- as_tibble(HairEyeColor) %>% 
  count(Hair, wt = n))
```

Fix the following bar plot so that it shows these counts.

```{r}
ggplot(hair) +
  aes(Hair, n) +
  geom_bar()
```

## Exercise 7: Tiling

Here's the number of people having a certain hair and eye colour from a sample of 592 people:

```{r}
(hair_eye <- as_tibble(HairEyeColor) %>% 
  count(Hair, Eye, wt = n))
```

Fix the following plot so that it shows a filled-in square for each combination. 
_Hint:_ What's the title of this exercise?

```{r}
ggplot(hair_eye, aes(Hair, Eye)) +
  geom_point(aes(colour = n))
```

By the way, `geom_count()` is like `geom_bar()`: it counts the number of overlapping points, but plots a point instead of a column.

## Additional take-home practice

If you'd like some practice, give these exercises a try

__Exercise 1__: 
Make a plot of `year` (x) vs `lifeExp` (y), with points colored by continent. 
Then, to that same plot, fit a **straight** regression line to each continent.

__Exercise 1b__: 
Repeat Exercise 1, but remove the error band from around the regression line.

__Exercise 2__: 
Repeat Exercise 1, but switch the _regression line_ and _geom\_point_ layers. 
How is this plot different from that of Exercise 1?

__Exercise 3__: 
Repeat Exercise 1, but omit the `geom_point()` layer. 
Does the line still show up, even though the data aren't shown? Why or why not?

__Exercise 4__:
Repeat Exercise 1, but add a **common regression line** that combines all continents.
Color this line **black**.
How does this common line differ from the continent-specific lines?

__Exercise 4__: 
Make a plot of `year` (x) vs `lifeExp` (y), faceted by continent. 
Then, fit a smoother through the data for each continent, without the error bars. Choose a span that you feel is appropriate.

__Exercise 5__: 
Plot the population over time (year) using lines, so that each country has its own line. 
Color by `gdpPercap`. 
Add alpha transparency to your liking.
Consider log-transforming population and GDP.

__Exercise 6__: 
Make a "raincloud" plot for life expectancy by continent.
For each continent, include (1) the individual data points, (2) a boxplot, and (3) a density curve.

**Save your finished document in the participation folder of your homework repo!**
