---
title: "Exercises: Multivariate pivoting"
output: 
  html_document:
    theme: paper
---

```{r}
library(tidyverse)
set.seed(123)
missing_w2_parent <- sample(1:500, 30)
missing_w2_child <- c(missing_w2_parent[1:5], sample(1:500, 25))
family  <- read_csv(
  "https://raw.githubusercontent.com/bwiernik/progdata/main/inst/tutorials/data/family_data.csv"
) %>% 
  mutate(
    across(
      starts_with("w2") & contains("parent"),
      ~ ifelse(family_id %in% missing_w2_parent, NA_real_, .x)
    ),
    across(
      starts_with("w2") & contains("child"),
      ~ ifelse(family_id %in% missing_w2_child, NA_real_, .x)
    )
  )
```

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```

You're working on a longitudinal study of parent-child relationships.
You have collected data from 500 families over 2 waves.
In each wave, both the child and parent completed measures of **communication behavior** and **relationship satisfaction**. 

```{r}
family %>% 
  DT::datatable(rownames = FALSE)
```

1. Reshape the dataset to a "longer" format.

  - Make each row **1 score**
  - Have columns for *family_id*, *family_member*, *wave*, *scale*, and *score*.

```{r}
family_longest <- family %>% 
  pivot_longer()

print(family_longest)
```

2. Reshape the dataset to a "longer" format.

  - Make each row **1 person**
  - Have columns for *family_id*, *family_member*, *wave*, *comm*, and *satis*.

```{r}
family_long <- family %>% 
  pivot_longer()

print(family_long)
```

3. Some families are missing wave 2 data for parent, child, or both. 
   Which families are missing wave 2 data for at least one person?

```{r}

```

Question: Is is easier to easier to find the missing data in the wide or long format?
