---
title: "Beer Advocate: ABV and quality ratings"
author: "Your Name Here"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this activity, download the data file at this location to the `data` subfolder in your `participation` folder. 
This dataset contains information about different catagories of beer as rated on <beeradvocate.com>.
The data are originally from [here](http://scottjanish.com/relationship-of-abv-to-beer-scores/).

```{r}
url <- "https://raw.githubusercontent.com/USF-Psych-DataSci/Classroom/master/data/beer-advocate.csv"
```

The dataset contains 3 variables:

  1. Beer category
  2. Average alcohol by volume (ABV) in percentage (%)
  3. Average quality rating (rAVG) on a 1–5 scale
  
You should write a script to do the following:

  1. Download the dataset with an appropriate filename.
  2. Import the dataset into R.
     Be sure to check for any importing problems and correct as needed.
  3. Make a figure illustrating the relationship between ABV and beer quality.
  4. Compute a summary table with the mean and SD of each variable, along with their correlation.
  5. Describe the results.
