---
title: 's12: Simulation Exercise'
output: 
  html_document:
    keep_md: true
    theme: paper
---

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)

#Sets our seed
set.seed(9999)
```

**When you make an Rmd file for participation or homework, be sure to do this**:

1. Change the file output to both html and md _documents_ (not notebook).
  - See the `keep_md: TRUE` argument above.

2. `knit` the document. 

3. Stage and commit the Rmd and knitted documents.


# Intro to simulations in R

Load the `tidyverse` package.
    - This will mainly be used for GGplot, but some of the data manipulation tools may be useful.

Hint: You can add the `suppressPackageStartupMessages()` function around the 
      `library()` command to keep your output looking nice!
    
```{r load packages, warning = FALSE, message = FALSE}
# load your packages here:
library(FILL_THIS_IN)
```
    

## `rnorm()`

1. Simulate two datasets that follow a normal distribution. The first one should have a mean of 4 and SD of 1. The second one should have a mean of 8 and SD of 1.5. Both should have 200 cases.

```{r}
dataNormal1 <- data.frame("Values" = FILL_THIS_IN,
                         #Creates a grouping factor for our data
                         "Group" = "Group1")

dataNormal2 <- data.frame("Values" = FILL_THIS_IN,
                         #Creates a grouping factor for our data
                         "Group" = "Group2")
```

2. Combine the two datasets.

```{r}
dataNormalCombined <- rbind(FILL_THIS_IN)
```

3. Graph these distributions using GGplot.

```{r}
ggplot(FILL_THIS_IN) +
  geom_histogram(alpha=0.6, position = 'identity')
```

4. Run summary stats of your three data sets

```{r}
summary(FILL_THIS_IN)
summary(FILL_THIS_IN)
summary(FILL_THIS_IN)
```

5. Run a t-test on your data

```{r}
t.test(dataNormalCombined$Values~dataNormalCombined$Group)
```
### Back to guide

## `sample()`

1. Sample from 5 numbers that follow an inverted normal distribution. (i.e. the probabilities are highest at the end points and lowest in the center).

```{r}

sample(c(FILL_THIS_IN), 200, replace=T, prob = c(FILL_THIS_IN)) 
hist(FILL_THIS_IN)

```

2. Pick one of the distributions that are [built into R](https://www.stat.umn.edu/geyer/old/5101/rlook.html) and replicate that distribution with the sample function. Make sure you graph both the original distribution and the reproduced distribution so you can make sure they match.

```{r}
#Add code for your initial distribution here. Make sure to use rDistributionFunction so that random numbers are generated.
FILL_THIS_IN
hist(FILL_THIS_IN)

#Replicate your distribution with the sample function.
sample(c(seq(FILL_THIS_IN)), #scale points
       10000, replace=T, prob = c(FILL_THIS_IN)) 
hist(FILL_THIS_IN)
```

### Back to guide

## ``for()`

1. Create a loop that generates a number from your favorite distribution and writes it to a column of your matrix. The matrix should have 20 rows and 1 column. You can use one of the distribution functions built into R, or create your own using `sample()`.
```{r}
matrix(FILL_THIS_IN)

for(FILL_THIS_IN)
{
  FILL_THIS_IN
}
hist(FILL_THIS_IN)
```

2. Write a loop within a loop. First, create a matrix with 20 rows and 6 columns. Save a number from your favorite distribution into each cell of the matrix. You can use one of the distribution functions built into R, or create your own using `sample()`. The distribution of responses should be the same for every person.

```{r}
matrix(FILL_THIS_IN)

for(FILL_THIS_IN)
{
  for (FILL_THIS_IN)
  {
   FILL_THIS_IN
  }
}

hist(FILL_THIS_IN)
```

3. Repeat 2, but this time make your distribution centered within each person.

```{r}
matrix(FILL_THIS_IN)

for(FILL_THIS_IN)
  {
  for(FILL_THIS_IN)
    {    
    if(FILL_THIS_IN
       {
         FILL_THIS_IN
       }
    else
      {
        FILL_THIS_IN
      }
    }
  }

hist(FILL_THIS_IN)
```

3. If needed, replace any out-of-bounds data.
```{r}
FILL_THIS_IN

hist(FILL_THIS_IN)
```

###Return to guide

## Conduct your own simulation study

You've now seen a few different methods we can use to simulate data. At this point you should be familiar with the `sample()`, `for()`, and `if()` functions and the distribution functions built into R. Now is your chance to use them to run your own simulation study.

1. Come up with an idea that can be simulated. 
- If you have a study idea you want to do a power analysis for this could be a good opportunity to try doing that
- You could simulate a study similar to our education intervention example. For instance, the implicit bias has been purported to influence police officer shooting decisions. One question might be, how big of a correlation would implicit bias need to have with police officer shooting decisions to account for 25% of the variance in those behaviors?
- There are also many papers that specify conceptual models of different psychological processes. You could try to replicate this conceptual model using simulated data.
2. Simulate your data. This simulation should use some of the more complex functions we covered, like a for loop, if statement, or some other type of logic. Just simulating one or two distributions using `rnorm()` is not enough for this part.
3. Write a summary of what you simulated and what you found.
NOTE: the resources section at the top of the main lab may be helpful if you're having trouble figuring out the right code to simulate what you want. This contains walkthroughs for different functions and simulation methods not covered here.
```{r}
FILL_THIS_IN
```