# Plotly

Consider the following plot:

```{r}
(p <- gapminder %>% 
     filter(continent != "Oceania") %>% 
     ggplot(aes(gdpPercap, lifeExp)) +
     geom_point(aes(colour=pop), alpha=0.2) +
     scale_x_log10(labels=dollar_format()) +
     scale_colour_distiller(
         trans   = "log10",
         breaks  = 10^(1:10),
         labels  = comma_format(),
         palette = "Greens"
     ) +
     facet_wrap(~ continent) +
     scale_y_continuous(breaks=10*(1:10)) +
     theme_bw())
```

1. Convert it to a `plotly` object by applying the `ggplotly()` function:

```{r}
FILL_THIS_IN
```

2. You can save a plotly graph locally as an html file. Try saving the above:
    - NOTE: plotly graphs don't seem to show up in Rmd _notebooks_, but they do with Rmd _documents_.

```{r}
p %>% 
    ggplotly() %>% 
    htmlwidgets::saveWidget("LOCATION_GOES_HERE")
```


3. Run this code to see the json format underneath:

```{r}
p %>% 
    ggplotly() %>% 
    plotly_json()
```


4. Check out code to make a plotly object from scratch using `plot_ly()` -- scatterplot of gdpPercap vs lifeExp.
    - Check out the [cheat sheet](https://images.plot.ly/plotly-documentation/images/r_cheat_sheet.pdf).

```{r}
plot_ly(gapminder, 
        x = ~gdpPercap, 
        y = ~lifeExp, 
        type = "scatter",
        mode = "markers",
        opacity = 0.2) %>% 
    layout(xaxis = list(type = "log"))
```

5. Add population to form a z-axis for a 3D plot:

```{r}
plot_ly(gapminder, 
        x = ~gdpPercap, 
        y = ~lifeExp, 
        z = FILL_THIS_IN,
        type = "scatter3d",
        mode = "markers",
        opacity = 0.2)
```
